---
title: "Working with Backends"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

joinspy works with base R data frames, tibbles, and data.tables. The join
wrappers (`left_join_spy()`, `join_strict()`, etc.) detect the input class and
dispatch to the right engine automatically. The diagnostic layer (`join_spy()`,
`key_check()`, `join_explain()`, and friends) is backend-agnostic: it runs
the same analysis regardless of what class the inputs are.

This vignette explains how detection works, when to override it, and what
happens to the output class through the diagnostic-repair-join cycle.


## Auto-detection

When you call `left_join_spy()` or `join_strict()` without specifying a
backend, joinspy inspects the class of `x` and `y` and picks the backend
according to a fixed priority:

1. If either input inherits from `data.table`, use data.table; if either inherits from `tbl_df` (tibble), use dplyr; otherwise fall back to base R `merge()`.

data.table wins over tibble because data.table's merge implementation has
specific expectations about key handling and indexing that cannot be replicated
through dplyr's join functions. A data.table joined via dplyr loses its keys,
its index, and its reference semantics, all properties that the caller likely
depends on if they chose data.table in the first place. dplyr, by contrast,
handles a data.table coerced to a tibble without issues. So when the two
classes conflict, data.table is the safer default. When both inputs are plain
data frames, base R is the only dependency-free option and is always available.

The detection examines both inputs, not just `x`. Pipelines sometimes pass a tibble on one side and a plain data frame on the other (for instance, reading one table from a CSV with `read.csv()` and another with `readr::read_csv()`). As long as one side is a tibble, dplyr is selected, which avoids the subtle differences in column ordering and NA handling between `merge()` and `dplyr::left_join()`.

What happens when a mixed-class call selects a backend whose package is not installed? joinspy falls back to base R and emits a warning. The most common trigger is a data.table object lingering in an environment where the data.table package was loaded transiently (e.g., via another package) but is no longer available. The warning is loud on purpose: a silent fallback could change the result's class and break downstream code.

Here is the detection in action with each input type:

```{r}
# Base R data frames: auto-detects "base"
orders_df <- data.frame(
  id = c(1, 2, 3),
  amount = c(100, 250, 75),
  stringsAsFactors = FALSE
)

customers_df <- data.frame(
  id = c(1, 2, 4),
  name = c("Alice", "Bob", "Diana"),
  stringsAsFactors = FALSE
)

result_base <- left_join_spy(orders_df, customers_df, by = "id", .quiet = TRUE)
class(result_base)
```

```{r eval = requireNamespace("dplyr", quietly = TRUE)}
# Tibbles: auto-detects "dplyr"
orders_tbl <- dplyr::tibble(
  id = c(1, 2, 3),
  amount = c(100, 250, 75)
)

customers_tbl <- dplyr::tibble(
  id = c(1, 2, 4),
  name = c("Alice", "Bob", "Diana")
)

result_dplyr <- left_join_spy(orders_tbl, customers_tbl, by = "id", .quiet = TRUE)
class(result_dplyr)
```

```{r eval = requireNamespace("data.table", quietly = TRUE)}
# data.tables: auto-detects "data.table"
orders_dt <- data.table::data.table(
  id = c(1, 2, 3),
  amount = c(100, 250, 75)
)

customers_dt <- data.table::data.table(
  id = c(1, 2, 4),
  name = c("Alice", "Bob", "Diana")
)

result_dt <- left_join_spy(orders_dt, customers_dt, by = "id", .quiet = TRUE)
class(result_dt)
```

When the two inputs have different classes, the higher-priority class wins:

```{r eval = requireNamespace("data.table", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)}
# data.table + tibble: data.table wins
mixed_result <- left_join_spy(orders_dt, customers_tbl, by = "id", .quiet = TRUE)
class(mixed_result)
```


## Explicit override

All join wrappers and `join_strict()` accept a `backend` argument that
overrides auto-detection. The three valid values are `"base"`, `"dplyr"`, and
`"data.table"`.

When would you use this? A few common scenarios:

**Forcing dplyr on plain data frames.** If you want tibble output (with its
stricter printing and no partial matching) but your inputs are plain data
frames, pass `backend = "dplyr"`:

```{r eval = requireNamespace("dplyr", quietly = TRUE)}
result <- left_join_spy(orders_df, customers_df, by = "id",
                        backend = "dplyr", .quiet = TRUE)
class(result)
```

**Forcing base R to avoid dplyr warnings.** dplyr 1.1.0 introduced a
many-to-many warning for joins that produce Cartesian products. If you have
already diagnosed the cardinality with `detect_cardinality()` or
`check_cartesian()` and know the expansion is intentional, you can bypass the
dplyr warning by forcing `backend = "base"`:

```{r eval = requireNamespace("dplyr", quietly = TRUE)}
# These have a legitimate many-to-many relationship
tags <- dplyr::tibble(
  item_id = c(1, 1, 2),
  tag = c("red", "large", "small")
)

prices <- dplyr::tibble(
  item_id = c(1, 2, 2),
  currency = c("USD", "USD", "EUR")
)

# Force base R to avoid dplyr's many-to-many warning
result <- left_join_spy(tags, prices, by = "item_id",
                        backend = "base", .quiet = TRUE)
nrow(result)
```

**Forcing data.table for speed on large data.** If you have plain data frames
with millions of rows, the data.table merge engine is substantially faster
than base R or dplyr. Passing `backend = "data.table"` triggers the conversion
internally; the result comes back as a data.table.

```{r eval = requireNamespace("data.table", quietly = TRUE)}
result <- left_join_spy(orders_df, customers_df, by = "id",
                        backend = "data.table", .quiet = TRUE)
class(result)
```

Note that the explicit backend must be installed. Requesting
`backend = "dplyr"` when dplyr is not installed will error, not silently fall
back. This is intentional: an explicit choice should fail loudly if it cannot
be honored. The asymmetry with auto-detection (which does fall back silently)
is deliberate. Auto-detection is a convenience; an explicit override is a
contract.

One more use case worth mentioning: **reproducibility across environments.**
If your script must produce identical output on a colleague's machine that may
or may not have dplyr installed, set `backend = "base"` explicitly. This
removes the class of the input as a variable and guarantees that the join
always uses `merge()`, regardless of whether someone upstream happened to
return a tibble or a plain data frame.


## Class preservation

joinspy preserves input class through the full diagnostic-repair-join cycle.
The three stages behave as follows:

- **Diagnostics** (`join_spy()`, `key_check()`, `join_explain()`, etc.)
  accept any data frame subclass and return report objects. They do not
  modify or convert the input data.
- **Repair** (`join_repair()`) operates on the key columns in place and
  returns the same class it received.
- **Join wrappers** (`left_join_spy()`, `join_strict()`, etc.) dispatch to
  the native join engine for the detected class, so the result matches the
  input class.

Here is a full cycle with base R data frames:

```{r}
messy_df <- data.frame(
  code = c("A-1 ", "B-2", " C-3"),
  value = c(10, 20, 30),
  stringsAsFactors = FALSE
)

lookup_df <- data.frame(
  code = c("A-1", "B-2", "C-3"),
  label = c("Alpha", "Beta", "Gamma"),
  stringsAsFactors = FALSE
)

# 1. Diagnose
report <- join_spy(messy_df, lookup_df, by = "code")

# 2. Repair
repaired_df <- join_repair(messy_df, by = "code")
class(repaired_df)  # still data.frame

# 3. Join
joined_df <- left_join_spy(repaired_df, lookup_df, by = "code", .quiet = TRUE)
class(joined_df)  # still data.frame
joined_df
```

The same cycle with tibbles:

```{r eval = requireNamespace("dplyr", quietly = TRUE)}
messy_tbl <- dplyr::tibble(
  code = c("A-1 ", "B-2", " C-3"),
  value = c(10, 20, 30)
)

lookup_tbl <- dplyr::tibble(
  code = c("A-1", "B-2", "C-3"),
  label = c("Alpha", "Beta", "Gamma")
)

repaired_tbl <- join_repair(messy_tbl, by = "code")
class(repaired_tbl)  # still tbl_df

joined_tbl <- left_join_spy(repaired_tbl, lookup_tbl, by = "code", .quiet = TRUE)
class(joined_tbl)  # still tbl_df
joined_tbl
```

And with data.tables:

```{r eval = requireNamespace("data.table", quietly = TRUE)}
messy_dt <- data.table::data.table(
  code = c("A-1 ", "B-2", " C-3"),
  value = c(10, 20, 30)
)

lookup_dt <- data.table::data.table(
  code = c("A-1", "B-2", "C-3"),
  label = c("Alpha", "Beta", "Gamma")
)

repaired_dt <- join_repair(messy_dt, by = "code")
class(repaired_dt)  # still data.table

joined_dt <- left_join_spy(repaired_dt, lookup_dt, by = "code", .quiet = TRUE)
class(joined_dt)  # still data.table
joined_dt
```

One edge case to be aware of: when you use `join_repair()` with both `x` and
`y` arguments, it returns a list with `$x` and `$y`. Each element preserves
the class of the corresponding input. If `x` is a tibble and `y` is a plain
data frame, you get a tibble in `$x` and a data frame in `$y`.

Class preservation also extends to `join_strict()`. The cardinality check
happens before the join, so if the constraint is satisfied the result has the
same class as the inputs, and if the constraint is violated no result is
returned at all (the function errors). There is no intermediate state where
the class might be lost.

The one place where class can change is when you use an explicit backend
override that does not match the input class. Passing `backend = "data.table"`
on a tibble will return a data.table, because the data.table merge engine
produces data.table output. Since you asked for the data.table engine, that is what you get. Pass `backend = "dplyr"` instead if you need a tibble back.


## Diagnostics are backend-agnostic

The diagnostic functions (`join_spy()`, `key_check()`, `key_duplicates()`,
`join_explain()`, `detect_cardinality()`, `check_cartesian()`) operate purely
on column values. They do not call any join engine. This means they produce
identical results whether the input is a data frame, tibble, or data.table.

This separation is useful in practice. You can diagnose with `join_spy()` on
a data.table, then join with dplyr if you prefer its output format. Or
diagnose in a script where only base R is available, then pass the data to a
Shiny app that uses dplyr internally. The diagnostic report does not carry any
backend dependency.

```{r eval = requireNamespace("data.table", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)}
# Diagnose on data.tables
orders_dt <- data.table::data.table(
  id = c(1, 2, 3),
  amount = c(100, 250, 75)
)

customers_dt <- data.table::data.table(
  id = c(1, 2, 4),
  name = c("Alice", "Bob", "Diana")
)

report <- join_spy(orders_dt, customers_dt, by = "id")

# Join with dplyr (convert first)
orders_tbl <- dplyr::as_tibble(orders_dt)
customers_tbl <- dplyr::as_tibble(customers_dt)
result <- left_join_spy(orders_tbl, customers_tbl, by = "id", .quiet = TRUE)
class(result)
```

The report from `join_spy()` is the same object regardless of input class. Its
`$issues`, `$expected_rows`, and `$match_analysis` fields contain the same
values; only the join execution step (called separately via the wrappers)
depends on the backend. This also means you can write unit tests for your key
quality using plain data frames even if your production code uses data.table.
The diagnostic output is structurally identical, so assertions on
`report$issues` or `report$match_analysis$match_rate` will hold regardless of
the class used at test time versus production time. You avoid pulling in heavy
Suggests dependencies just to test key diagnostics.


## Which backend to choose

joinspy does not have a preferred backend. It respects whatever you are
already using. Here is a brief orientation for cases where you do need to
choose.

**Base R** has no dependencies and is always available. It uses `merge()`,
which handles all four join types and named `by` vectors for different column
names. The output is always a plain data frame. Row order is not guaranteed
(it depends on the key sort), and `merge()` reorders columns to put the key
first. For small to medium datasets (up to a few hundred thousand rows), base
R is perfectly adequate.

**dplyr** is the standard in tidyverse workflows. It preserves column order
and row order from the left table, returns tibbles, and prints compactly. It
requires the dplyr package (and transitively tibble and several others). Since
dplyr 1.1.0, it also warns about many-to-many joins, which is helpful as a
safety net but can be noisy if you have already diagnosed the cardinality with
joinspy.

**data.table** is the fastest option for large datasets. Its merge
implementation is heavily optimized and can handle tens of millions of rows
efficiently. The tradeoff is reference semantics: data.table modifies objects
in place in some operations, which can surprise users accustomed to R's
copy-on-modify behavior. joinspy's data.table backend uses `merge()` (not the
`[.data.table` join syntax), so it avoids most reference-semantic surprises,
but the result is still a data.table object with data.table's printing and
subsetting behavior.

A practical consideration: the three backends differ in how they handle column
name collisions. When `x` and `y` share non-key column names, base R appends
`.x` and `.y` suffixes, dplyr appends `.x` and `.y` by default (configurable
via `suffix`), and data.table appends `i.` to the columns from the right table.
These suffixes affect downstream code that references the merged columns by
name. If you switch backends mid-project, check that column references still
resolve correctly.

Another difference is row ordering. Base R's `merge()` sorts rows by the key
column. dplyr preserves the row order of the left table. data.table sorts by
key if the tables are keyed, and otherwise preserves insertion order. For most
analytical workflows the order does not matter, but if you are comparing output
across backends in tests or reproducibility checks, sort the result before
comparing.

The simplest rule: use what your project already depends on. If you have no
preference, base R is the safest default. If performance matters, data.table.
If you want tibble output and tidyverse compatibility, dplyr.


## See Also

- `vignette("quickstart")` for a quick introduction to joinspy

- `vignette("common-issues")` for a catalogue of join problems and solutions

- `?left_join_spy`, `?join_strict` for backend parameter documentation
