---
title: "Common Join Problems"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Common Join Problems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

Most join bugs are not about cardinality. They are about strings. A trailing
space, a flipped case, a zero-width Unicode character: any of these will cause
two keys that *look* identical to fail matching without a warning from R.

This vignette walks through the problems joinspy catches, ordered by
frequency. String-level issues come first because they account for the
majority of "why did my join drop rows?" questions.
Structural issues (duplicates, NAs, type mismatches, Cartesian explosions)
follow.

## Trailing and leading whitespace

The classic. Someone exports a CSV from Excel, and now half your keys carry a
trailing space. Everything looks fine when you print the data frame. Nothing
matches when you join.

```{r}
sales <- data.frame(
  product = c("Widget", "Gadget ", " Gizmo"),
  units = c(10, 20, 30),
  stringsAsFactors = FALSE
)

inventory <- data.frame(
  product = c("Widget", "Gadget", "Gizmo"),
  stock = c(100, 200, 300),
  stringsAsFactors = FALSE
)

join_spy(sales, inventory, by = "product")
```

Two of three keys carry whitespace that prevents matching. `join_repair()`
strips it:

```{r}
sales_clean <- join_repair(sales, by = "product")
key_check(sales_clean, inventory, by = "product")
```

If you want to see the repair without applying it, pass `dry_run = TRUE`.

Why does this happen so often? Excel pads cells when you widen columns manually,
and copy-pasting from web tables brings in non-breaking spaces that look like
regular ones. ETL pipelines that concatenate fixed-width fields frequently leave
padding intact, and the engineer who wrote the pipeline assumed downstream
consumers would trim—an assumption that rarely survives contact with reality.

The problem compounds with composite keys. When the join uses two or more
columns, whitespace in *any* of them is enough to break the match:

```{r}
shipments <- data.frame(
  warehouse = c("East ", "West", "East "),
  product   = c("Widget", "Gadget ", "Gizmo"),
  shipped   = c(50, 80, 35),
  stringsAsFactors = FALSE
)

stock <- data.frame(
  warehouse = c("East", "West", "East"),
  product   = c("Widget", "Gadget", "Gizmo"),
  on_hand   = c(200, 150, 90),
  stringsAsFactors = FALSE
)

join_spy(shipments, stock, by = c("warehouse", "product"))
```

Both `warehouse` and `product` carry trailing spaces in `shipments`, so all
three rows fail to match despite looking correct on screen. A single
`join_repair()` call cleans every key column at once:

```{r}
shipments_clean <- join_repair(shipments, by = c("warehouse", "product"))
key_check(shipments_clean, stock, by = c("warehouse", "product"))
```

## Case mismatches

Databases are often case-insensitive. R is not. When you pull tables from two
different systems, one might store `"ABC"` and the other `"abc"`. They refer to
the same entity, but R treats them as different keys.

```{r}
sensors <- data.frame(
  station = c("AWS-01", "aws-02", "Aws-03"),
  temp = c(22.1, 18.4, 25.7),
  stringsAsFactors = FALSE
)

metadata <- data.frame(
  station = c("aws-01", "AWS-02", "AWS-03"),
  region = c("North", "South", "East"),
  stringsAsFactors = FALSE
)
```

None of these keys match as-is. `join_spy()` picks this up:

```{r}
join_spy(sensors, metadata, by = "station")
```

Repair both sides to a common case:

```{r}
repaired <- join_repair(sensors, metadata, by = "station", standardize_case = "lower")
key_check(repaired$x, repaired$y, by = "station")
```

You choose the target case. `"lower"`, `"upper"`, or `"title"` all work.
In practice, `"lower"` is the safest default. Lowercase is unambiguous (no
locale-dependent rules for title case), and most database systems normalize to
lowercase internally. Upper case is fine too, but title case introduces edge
cases with acronyms and prepositions that vary by language.

One subtlety: if your key column is a `factor`, case standardization changes the
level labels but not the underlying integer codes. This can produce factors whose
levels look identical after lowering but originally pointed to different groups.
Convert to character first (`as.character()`) before standardizing case on factor
columns; otherwise you risk silently merging distinct factor levels.

## Encoding and invisible characters

A key contains a non-breaking space (U+00A0) instead of a regular space, or
a zero-width joiner crept in during a copy-paste from a PDF. The strings
print identically. They do not match.

```{r}
# Simulate a non-breaking space in one key
left <- data.frame(
  city = c("New York", "Los\u00a0Angeles", "Chicago"),
  pop = c(8.3, 3.9, 2.7),
  stringsAsFactors = FALSE
)

right <- data.frame(
  city = c("New York", "Los Angeles", "Chicago"),
  area = c(302, 469, 227),
  stringsAsFactors = FALSE
)

join_spy(left, right, by = "city")
```

The `"Los\u00a0Angeles"` key in `left` looks like `"Los Angeles"` in `right`,
but the non-breaking space makes them different byte sequences.
`join_repair()` with `remove_invisible = TRUE` (the default) strips these out:

```{r}
left_fixed <- join_repair(left, by = "city")
key_check(left_fixed, right, by = "city")
```

Where do these invisible characters come from? The most common sources are:
PDF extraction (which inserts soft hyphens, zero-width spaces, and ligature
code points that have no visible glyph); web scraping, where HTML entities and
Unicode normalization forms vary across pages; and cross-platform file transfers,
where Windows line endings (`\r\n`) or UTF-8 BOM markers (the byte sequence
`EF BB BF` at the start of a file) sneak into key columns when files are read
as raw text. Spreadsheets saved from Google Sheets vs. Excel vs. LibreOffice
produce subtly different Unicode normalization, so the same accented character
may be stored as a single code point in one file and as a base character plus a
combining accent in another.

`join_repair()` with `remove_invisible = TRUE` (the default) handles the most
common offenders: non-breaking spaces, zero-width joiners and non-joiners, BOM
markers, and soft hyphens. It does not attempt full Unicode normalization (NFC
vs. NFD), nor does it strip all control characters; some of those are legitimate
data. If `join_spy()` still flags encoding issues after repair, you may need
`stringi::stri_trans_nfc()` for normalization or a manual inspection of the raw
byte sequences with `chartr()`.

When you suspect encoding trouble but cannot see it, `suggest_repairs()` on a
report will print the exact R code you need.

## Empty strings masquerading as data

Empty strings (`""`) are valid character values in R. They will match other
empty strings in a join, which is almost never what you want: two rows with
missing identifiers get joined as though they refer to the same entity.

```{r}
patients <- data.frame(
  mrn = c("P001", "", "P003"),
  age = c(34, 56, 29),
  stringsAsFactors = FALSE
)

visits <- data.frame(
  mrn = c("P001", "P002", ""),
  date = c("2024-01-10", "2024-02-15", "2024-03-20"),
  stringsAsFactors = FALSE
)

join_spy(patients, visits, by = "mrn")
```

Convert empties to `NA` before joining. NAs never match in R, which is the
correct behavior for missing identifiers:

```{r}
patients_fixed <- join_repair(patients, by = "mrn", empty_to_na = TRUE)
patients_fixed$mrn
```

The join will now drop those rows instead of incorrectly linking them, but
note that `data.table` treats `""` and `NA_character_` as distinct values in
keyed joins; an empty string in one table will not match an `NA` in the other,
even though both represent "missing." If you use a data.table backend, convert
empty strings to `NA` on both sides to get consistent behavior.

## Near-matches and typos

Sometimes keys are close but not identical. `"Wiget"` vs `"Widget"`, or
`"customer_001"` vs `"cust_001"`. These are not whitespace or case problems;
they are genuine mismatches that `join_spy()` can flag when it finds keys in
one table with no counterpart in the other.

```{r}
orders <- data.frame(
  sku = c("WDG-100", "GDG-200", "GZM-300"),
  qty = c(5, 12, 8),
  stringsAsFactors = FALSE
)

catalog <- data.frame(
  sku = c("WDG-100", "GDG-200", "GZM-301"),
  price = c(9.99, 14.99, 7.50),
  stringsAsFactors = FALSE
)

report <- join_spy(orders, catalog, by = "sku")
```

The mismatch between `"GZM-300"` and `"GZM-301"` shows up in the unmatched
keys. Internally, `join_spy()` computes edit distances (Levenshtein distance)
between unmatched keys on each side. When two keys differ by only one or two
characters, the report flags them as near-matches, which helps you distinguish
typos from genuinely different entities.

Here is a clearer example where near-match detection fires on multiple keys:

```{r}
employees <- data.frame(
  name = c("Johnson", "Smithe", "O'Brian", "Williams"),
  dept = c("Sales", "R&D", "Ops", "HR"),
  stringsAsFactors = FALSE
)

payroll <- data.frame(
  name = c("Jonhson", "Smith", "O'Brien", "Williams"),
  salary = c(55000, 62000, 48000, 71000),
  stringsAsFactors = FALSE
)

report <- join_spy(employees, payroll, by = "name")
```

`"Johnson"` vs. `"Jonhson"` (transposition), `"Smithe"` vs. `"Smith"` (extra
character), and `"O'Brian"` vs. `"O'Brien"` (vowel swap) are all within edit
distance 2 and get flagged. `"Williams"` matches exactly and passes through.

For typos, there is no automated fix because joinspy cannot know which
side is correct. But the near-match list gives you a concrete starting point:
build a lookup table mapping the wrong spellings to the right ones, or go back
to the source system and fix the data there.

## Duplicate keys

Now we cross into structural territory. Duplicate keys cause row
multiplication. A left join on a key that appears twice in the right table will
double the corresponding rows from the left table. Three copies triple them.
Most people do not expect this.

```{r}
orders <- data.frame(
  customer_id = c(1, 2, 3),
  amount = c(100, 250, 75)
)

addresses <- data.frame(
  customer_id = c(1, 2, 2, 3),
  address = c("NYC", "LA", "SF", "Chicago"),
  stringsAsFactors = FALSE
)

join_spy(orders, addresses, by = "customer_id")
```

`key_duplicates()` shows you exactly which rows are responsible:

```{r}
key_duplicates(addresses, by = "customer_id")
```

The fix depends on your intent. If each customer should have one address,
deduplicate first. If you genuinely need all combinations, the multiplication
is correct and you just need to know it will happen.

When *both* sides have duplicates the situation is worse. Each key group
produces a Cartesian product:

```{r}
orders_dup <- data.frame(
  product = c("A", "A", "B", "B"),
  qty     = c(10, 20, 5, 15)
)

prices_dup <- data.frame(
  product = c("A", "A", "A", "B", "B"),
  price   = c(1.0, 1.1, 1.2, 2.0, 2.5)
)

join_spy(orders_dup, prices_dup, by = "product")
```

Product `"A"` has 2 rows on the left and 3 on the right, so a join produces
2 x 3 = 6 rows for that key alone. `check_cartesian()` quantifies the total
expansion across all key groups before you run the join:

```{r}
check_cartesian(orders_dup, prices_dup, by = "product")
```

## NA keys

`NA` never equals `NA` in R. This is by design, but it surprises people who
expect two missing values to match.

```{r}
orders <- data.frame(
  customer_id = c(1, NA, 3, NA),
  amount = c(100, 200, 300, 400)
)

customers <- data.frame(
  customer_id = c(1, 2, 3, NA),
  name = c("Alice", "Bob", "Carol", "Unknown"),
  stringsAsFactors = FALSE
)

join_spy(orders, customers, by = "customer_id")
```

NA keys are common in datasets where some records were entered without a
complete identifier (think hospital admissions with unknown patient IDs, or
sensor readings before the device was registered). The deeper problem is that
NA keys don't just fail to match the other table; they also fail to match
*each other*. Two rows with `NA` keys in the left table will not be grouped
together, which means aggregation after a join treats each NA row as its own
group. This is rarely what you want.

There are two reasonable responses. Remove rows with NA keys before joining, or
replace NAs with a sentinel value if you actually want them to match:

```{r}
# Remove
orders_clean <- orders[!is.na(orders$customer_id), ]
key_check(orders_clean, customers, by = "customer_id")
```

## Type mismatches

One table stores IDs as integers, the other as character strings.
`merge()` will coerce silently; `dplyr::left_join()` will refuse. Either way,
you want to know about it before the join, not after.

```{r}
invoices <- data.frame(
  product_id = c(1, 2, 3),
  total = c(500, 300, 150)
)

products <- data.frame(
  product_id = c("1", "2", "3"),
  name = c("Widget", "Gadget", "Gizmo"),
  stringsAsFactors = FALSE
)

join_spy(invoices, products, by = "product_id")
```

A subtler variant occurs with `Date` vs. character, or `POSIXct` vs. `Date`:
the join either fails outright or coerces through numeric intermediaries, producing
nonsense matches. `join_spy()` flags the type mismatch regardless of which types
are involved, not just integer-vs-character.

The fix is always explicit conversion. Pick the type that makes more sense for
your downstream analysis and convert the other table:

```{r}
invoices$product_id <- as.character(invoices$product_id)
key_check(invoices, products, by = "product_id")
```

## Many-to-many explosions

When both tables have duplicate keys, you get a Cartesian product within each
key group. Two duplicates on the left times three on the right gives six rows
for that key alone. With real data this can turn a 10,000-row join into a
million-row table. To put concrete numbers on it: suppose the left table has
50,000 rows and the right table has 30,000 rows, with an average of 4
duplicates per key on each side. The resulting join can exceed 100 million rows,
consuming gigabytes of memory and taking minutes to complete on hardware that
handles the original tables in milliseconds.

```{r}
items <- data.frame(
  order_id = c(1, 1, 2, 2, 2),
  item = c("A", "B", "C", "D", "E"),
  stringsAsFactors = FALSE
)

payments <- data.frame(
  order_id = c(1, 1, 2, 2),
  method = c("Card", "Cash", "Card", "Wire"),
  stringsAsFactors = FALSE
)

check_cartesian(items, payments, by = "order_id")
```

`detect_cardinality()` tells you the relationship type:

```{r}
detect_cardinality(items, payments, by = "order_id")
```

In production, m:m joins are a common source of wrong aggregations. A
50,000-row table joined against a 30,000-row table with moderate duplication
can produce tens of millions of rows. Downstream sums, means, and counts then
operate on the inflated table and return incorrect numbers, because
observations are counted multiple times.

If you expected a one-to-many relationship, `join_strict()` will stop you
before the explosion happens:

```{r error = TRUE}
join_strict(items, payments, by = "order_id", type = "left", expect = "1:m")
```

## No matches at all

Your inner join returns zero rows. The result is an empty data frame, which
downstream code may not check for. The first thing to verify is whether the
keys overlap at all:

```{r}
system_a <- data.frame(
  user_id = c("USR-001", "USR-002", "USR-003"),
  score = c(85, 90, 78),
  stringsAsFactors = FALSE
)

system_b <- data.frame(
  user_id = c("1", "2", "3"),
  dept = c("Sales", "R&D", "Ops"),
  stringsAsFactors = FALSE
)

join_spy(system_a, system_b, by = "user_id")
```

Zero overlap. The keys use completely different formats. No amount of trimming
or case-folding will help here; the problem is semantic, not syntactic. This
happens frequently when two systems were designed independently, or when one
system migrated its ID scheme and the other did not. You need a mapping table
or a transformation that extracts the numeric part from `system_a`:

```{r}
system_a$user_num <- gsub("^USR-0*", "", system_a$user_id)
key_check(system_a, system_b, by = c("user_num" = "user_id"))
```

## Troubleshooting workflow

When a join goes wrong, work through these steps in order. Most problems
resolve in the first three.

**Step 1: Run `join_spy()`.** This is always the starting point. It checks
string quality, key overlap, cardinality, and predicted row counts in a single
call.

```{r eval = FALSE}
report <- join_spy(x, y, by = "key_col")
```

For large datasets (hundreds of thousands of rows or more), the full diagnostic
scan can be slow. Pass `sample = 1000` to run `join_spy()` on a random subset
first; this catches the most common problems in seconds rather than minutes. If
the sampled report looks clean, run again without sampling to confirm.

**Step 2: Fix string issues.** If the report flags whitespace, case, encoding,
or empty strings, run `join_repair()`. This handles the most common class of
silent failures.

```{r eval = FALSE}
x_clean <- join_repair(x, by = "key_col")
# or both sides:
repaired <- join_repair(x, y, by = "key_col", standardize_case = "lower")
```

**Step 3: Check for duplicates.** If the predicted row count is higher than you
expected, inspect duplicates with `key_duplicates()` and decide whether to
aggregate or deduplicate.

```{r eval = FALSE}
key_duplicates(y, by = "key_col")
```

**Step 4: Enforce cardinality.** Once keys are clean, use `join_strict()` to
guarantee the relationship you expect. This catches edge cases that slip
through manual inspection.

```{r eval = FALSE}
result <- join_strict(x_clean, y_clean, by = "key_col",
                      type = "left", expect = "1:1")
```

**Step 5: Explain the result.** If the joined output still looks wrong, run
`join_explain()` to get a breakdown of what happened:

```{r eval = FALSE}
result <- left_join_spy(x_clean, y_clean, by = "key_col")
join_explain(result, x_clean, y_clean, by = "key_col")
```

**Step 6: Log for audit trails.** In production workflows, you want a record of
what `join_spy()` found and what repairs were applied. Use `set_log_file()` at
the top of your pipeline to route all subsequent reports to a log file
automatically:

```{r eval = FALSE}
set_log_file("logs/join_diagnostics.log")
# All join_spy() / join_explain() calls now append to this file
```

This is especially useful when debugging issues days or weeks after the pipeline
ran; the log tells you exactly what the key distributions looked like at join
time.

The pattern is: diagnose strings first, then structure, then cardinality.
String problems are invisible and common. Structural problems are visible once
you know to look. If you find yourself repeating this workflow across multiple
joins in the same pipeline, consider using `analyze_join_chain()` to diagnose an
entire sequence of joins at once; it applies the same checks to every step and
reports where the first problem enters the chain.

## See Also

- `vignette("quickstart")` for a quick introduction to joinspy

- `?join_spy`, `?join_repair`, `?key_check`, `?join_strict`

- `?check_cartesian`, `?detect_cardinality`, `?join_explain`
