---
title: "Common Join Problems"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Common Join Problems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

## Overview

This vignette catalogs common join problems and shows how joinspy detects and resolves them. Each issue includes detection methods and recommended solutions.

## Issue 1: Duplicate Keys

**Problem**: When one or both tables have duplicate keys, joins multiply rows unexpectedly.

```{r}
orders <- data.frame(
  customer_id = c(1, 2, 2, 3),
  amount = c(100, 50, 75, 200)
)

addresses <- data.frame(
  customer_id = c(1, 2, 2, 3),
  address = c("NYC", "LA", "SF", "Chicago")
)

join_spy(orders, addresses, by = "customer_id")
```

**Detection**: `join_spy()` reports duplicate counts and expected row multiplication.

**Solution**: Aggregate or filter duplicates before joining.

```{r}
key_duplicates(orders, by = "customer_id")
key_duplicates(addresses, by = "customer_id")
```

## Issue 2: Whitespace

**Problem**: Invisible leading/trailing spaces prevent matches.

```{r}
sales <- data.frame(
  product = c("Widget", "Gadget ", " Gizmo"),
  units = c(10, 20, 30),
  stringsAsFactors = FALSE
)

inventory <- data.frame(
  product = c("Widget", "Gadget", "Gizmo"),
  stock = c(100, 200, 300),
  stringsAsFactors = FALSE
)

join_spy(sales, inventory, by = "product")
```

**Detection**: `join_spy()` flags whitespace issues in the Issues section.

**Solution**: Use `join_repair()` or `trimws()`.

```{r}
sales_fixed <- join_repair(sales, by = "product")
key_check(sales_fixed, inventory, by = "product")
```

## Issue 3: Case Mismatches

**Problem**: Keys differ only by case ("ABC" vs "abc").

```{r}
left <- data.frame(
  code = c("ABC", "def", "GHI"),
  value = 1:3,
  stringsAsFactors = FALSE
)

right <- data.frame(
  code = c("abc", "DEF", "ghi"),
  label = c("A", "D", "G"),
  stringsAsFactors = FALSE
)

join_spy(left, right, by = "code")
```

**Detection**: `join_spy()` detects case mismatches when keys would match if case-insensitive.

**Solution**: Standardize case with `join_repair()`.

```{r}
repaired <- join_repair(left, right, by = "code", standardize_case = "upper")
key_check(repaired$x, repaired$y, by = "code")
```

## Issue 4: NA Keys

**Problem**: NA values in key columns never match (by design in R).

```{r}
orders <- data.frame(
  customer_id = c(1, NA, 3, NA),
  amount = c(100, 200, 300, 400)
)

customers <- data.frame(
  customer_id = c(1, 2, 3, NA),
  name = c("Alice", "Bob", "Carol", "Unknown")
)

join_spy(orders, customers, by = "customer_id")
```

**Detection**: `join_spy()` reports NA counts in the Table Summary.

**Solution**: Handle NA values explicitly—remove them or replace with a placeholder.

```{r}
# Option 1: Remove rows with NA keys
orders_clean <- orders[!is.na(orders$customer_id), ]

# Option 2: Replace NA with placeholder
orders$customer_id[is.na(orders$customer_id)] <- -999
```

## Issue 5: No Matches

**Problem**: Inner join returns zero rows when you expected matches.

```{r}
system_a <- data.frame(
  user_id = c("USR001", "USR002", "USR003"),
  score = c(85, 90, 78),
  stringsAsFactors = FALSE
)

system_b <- data.frame(
  user_id = c("1", "2", "3"),
  department = c("Sales", "Marketing", "Engineering"),
  stringsAsFactors = FALSE
)

report <- join_spy(system_a, system_b, by = "user_id")
```

**Detection**: Match analysis shows 0% match rate.

**Solution**: Create a mapping table or transform keys to a common format.

```{r}
# Extract numeric part
system_a$user_num <- gsub("USR0*", "", system_a$user_id)
key_check(system_a, system_b, by = c("user_num" = "user_id"))
```

## Issue 6: Many-to-Many Explosion

**Problem**: Both tables have duplicate keys, causing exponential row growth.

```{r}
order_items <- data.frame(
  order_id = c(1, 1, 2, 2, 2),
  item = c("A", "B", "C", "D", "E")
)

order_payments <- data.frame(
  order_id = c(1, 1, 2, 2),
  payment = c("CC1", "CC2", "Cash", "Check")
)

report <- join_spy(order_items, order_payments, by = "order_id")
```

**Detection**: Expected row counts show multiplication (inner join = 10 rows from 9 source rows).

**Solution**: Aggregate one table first, or use `check_cartesian()`.

```{r}
check_cartesian(order_items, order_payments, by = "order_id")
```

```{r error = TRUE}
# Enforce cardinality to catch this
join_strict(order_items, order_payments, by = "order_id", expect = "1:m")
```

## Issue 7: Type Mismatches

**Problem**: Keys have different types (numeric vs character).
```{r}
orders <- data.frame(
  product_id = c(1, 2, 3),
  quantity = c(10, 20, 30)
)

products <- data.frame(
  product_id = c("1", "2", "3"),
  name = c("Widget", "Gadget", "Gizmo"),
  stringsAsFactors = FALSE
)

join_spy(orders, products, by = "product_id")
```

**Detection**: `join_spy()` flags type coercion warnings.

**Solution**: Convert to matching types before joining.

```{r}
orders$product_id <- as.character(orders$product_id)
key_check(orders, products, by = "product_id")
```

## Issue 8: Empty Strings vs NA

**Problem**: Empty strings (`""`) and `NA` behave differently in joins.

```{r}
left <- data.frame(
  id = c("A", "", "C"),
  value = 1:3,
  stringsAsFactors = FALSE
)

right <- data.frame(
  id = c("A", "B", ""),
  label = c("Alpha", "Beta", "Empty"),
  stringsAsFactors = FALSE
)

join_spy(left, right, by = "id")
```

**Detection**: `join_spy()` warns about empty strings in keys.

**Solution**: Convert empty strings to NA with `join_repair()`.

```{r}
left_fixed <- join_repair(left, by = "id", empty_to_na = TRUE)
left_fixed$id
```

## Using join_strict() for Safety

When you know the expected cardinality, use `join_strict()` to fail fast:

```{r error = TRUE}
products <- data.frame(id = 1:3, name = c("A", "B", "C"))
prices <- data.frame(id = c(1, 2, 2, 3), price = c(10, 20, 25, 30))

join_strict(products, prices, by = "id", expect = "1:1")
```

## Automatic Detection with detect_cardinality()

Let joinspy determine the actual relationship:

```{r}
orders <- data.frame(id = c(1, 1, 2, 3), item = c("A", "B", "C", "D"))
customers <- data.frame(id = 1:3, name = c("Alice", "Bob", "Carol"))

detect_cardinality(orders, customers, by = "id")
```

## Quick Reference

| Issue | Detection | Solution |
|-------|-----------|----------|
| Duplicates | `join_spy()`, `key_duplicates()` | Aggregate or filter |
| Whitespace | `join_spy()`, `key_check()` | `join_repair()`, `trimws()` |
| Case mismatch | `join_spy()` | `join_repair(standardize_case=)` |
| NA keys | `join_spy()` Table Summary | Remove or replace |
| No matches | `join_spy()` Match Analysis | Check key format/mapping |
| M:M explosion | `join_spy()`, `check_cartesian()` | Aggregate first |
| Type mismatch | `join_spy()` | Convert types |
| Empty strings | `join_spy()` | `join_repair(empty_to_na=TRUE)` |

## Troubleshooting Workflow

1. Run `join_spy(x, y, by)` to get a comprehensive diagnostic
2. Check the Issues section for detected problems
3. Use `key_duplicates()` to locate specific duplicate rows
4. Apply `join_repair()` to fix whitespace/case/encoding issues
5. Use `join_strict()` to enforce expected cardinality
6. After joining, use `join_explain()` to understand row count changes

## See Also

- `vignette("quickstart")` - Quick start guide
- `?join_spy`, `?key_check`, `?join_repair`, `?join_strict`
- `?check_cartesian`, `?detect_cardinality`

## Session Info

```{r}
sessionInfo()
```
