---
title: "Joins in Production"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joins in Production}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

Interactive analysis and production scripts have different failure modes. When
you explore data at the console, a bad join is annoying but recoverable: you
see the wrong row count, swear, fix the key, and re-run. In a scheduled script
that feeds a dashboard or triggers downstream jobs, that same bad join silently
corrupts everything it touches. Nobody notices until a stakeholder files a
ticket three weeks later.

joinspy was built with both contexts in mind, but this vignette focuses
exclusively on the second one: wiring joinspy into automated pipelines so
that join problems surface as errors or log entries rather than silent data
corruption. Every example here uses synthetic data and `tempfile()` paths, so
you can run the vignette end to end without touching your filesystem.

## Assertions with key_check()

`key_check()` returns a single logical: `TRUE` if no issues were detected,
`FALSE` otherwise. That makes it a natural fit for assertions. The simplest
pattern wraps it in `stopifnot()`:

```{r}
orders <- data.frame(
  customer_id = c("C01", "C02", "C03", "C04"),
  amount = c(100, 200, 150, 300),
  stringsAsFactors = FALSE
)

customers <- data.frame(
  customer_id = c("C01", "C02", "C03", "C04"),
  region = c("East", "West", "East", "North"),
  stringsAsFactors = FALSE
)

# This passes -- keys are clean
stopifnot(key_check(orders, customers, by = "customer_id", warn = FALSE))
```

When the keys have problems, the script halts:

```{r error = TRUE}
orders_dirty <- data.frame(
  customer_id = c("C01", "C02 ", "C03 ", "C04"),
  amount = c(100, 200, 150, 300),
  stringsAsFactors = FALSE
)

stopifnot(key_check(orders_dirty, customers, by = "customer_id", warn = FALSE))
```

The `warn = FALSE` argument suppresses the printed diagnostics. In a cron job
or CI pipeline, you generally want the script to fail hard and fast rather than
print warnings that nobody reads. If you need more control over the error
message, use an explicit `if`/`stop` pattern instead:

```{r error = TRUE}
if (!key_check(orders_dirty, customers, by = "customer_id", warn = FALSE)) {
  stop("Key quality check failed for orders-customers join. ",
       "Run join_spy() interactively for details.", call. = FALSE)
}
```

This is more verbose but gives you a custom error message that can include the
table names, the pipeline step, or a link to documentation. You choose the
tradeoff.

A common pattern chains the assertion with a repair step. Run `key_check()`
first; if it fails, repair and re-check:

```{r}
ok <- key_check(orders_dirty, customers, by = "customer_id", warn = FALSE)

if (!ok) {
  repaired <- join_repair(
    orders_dirty, customers,
    by = "customer_id",
    trim_whitespace = TRUE,
    remove_invisible = TRUE
  )
  orders_clean <- repaired$x
  customers_clean <- repaired$y

  # Re-check after repair
  stopifnot(key_check(orders_clean, customers_clean,
                       by = "customer_id", warn = FALSE))
}
```

When should you use `key_check()` versus `join_spy()`? They serve different
roles. `key_check()` is a gate, a binary pass/fail that costs almost nothing
to run. `join_spy()` is a detailed diagnostic that builds a full report
object with match rates, expected row counts, and categorized issues. In a
production script, `key_check()` is the assertion you run on every execution;
`join_spy()` is what you pull out when the assertion fails and you need to
understand why. Calling `join_spy()` on every run is fine for moderate-sized
tables, but for million-row datasets the overhead adds up, and `key_check()`
is the leaner choice.

## Silent Joins in Pipelines

The `*_join_spy()` wrappers print diagnostic output by default. That output is
helpful when you are working interactively; it is noise in a scheduled script.
The `.quiet = TRUE` argument suppresses all printed output while still
computing and storing the diagnostic report internally.

```{r}
sensors <- data.frame(
  sensor_id = c("S01", "S02", "S03", "S04"),
  location = c("Roof", "Basement", "Lobby", "Garage"),
  stringsAsFactors = FALSE
)

readings <- data.frame(
  sensor_id = c("S01", "S02", "S03", "S05"),
  temperature = c(22.1, 18.5, 21.0, 19.3),
  stringsAsFactors = FALSE
)

# Nothing printed
result <- left_join_spy(sensors, readings, by = "sensor_id", .quiet = TRUE)
```

The report is still available after the fact via `last_report()`:

```{r}
rpt <- last_report()
rpt$match_analysis$match_rate
```

This separation of execution and inspection is the key pattern. The join runs
silently inside a pipeline. Later (maybe in a validation step at the end of
the script, maybe in a separate monitoring job) you pull the report and check
its contents programmatically:

```{r}
rpt <- last_report()
if (rpt$match_analysis$match_rate < 0.95) {
  warning(sprintf(
    "Low match rate (%.1f%%) in sensor join -- check for missing sensor IDs",
    rpt$match_analysis$match_rate * 100
  ))
}
```

This gives you the diagnostic power of `join_spy()` without cluttering the
standard output of your pipeline. The report object persists in memory until the
next `*_join_spy()` call overwrites it, so there is no race condition as long
as your script is single-threaded (which R scripts are).

You can build arbitrarily complex validation logic on top of the report object.
Check the number of issues, inspect specific issue types, compare the expected
row count against a threshold; whatever your pipeline requires. The report is
a plain list, so standard R subsetting works. Nothing about the validation
requires interactive use; it is all programmatic.

One caveat: `last_report()` only stores the single most recent report. If your
script performs three joins in sequence, only the third report survives. If you
need to retain reports from earlier joins, either log them (see the next
section) or capture them explicitly:

```{r}
result1 <- left_join_spy(sensors, readings, by = "sensor_id", .quiet = TRUE)
report1 <- last_report()

# ... later ...
result2 <- inner_join_spy(sensors, readings, by = "sensor_id", .quiet = TRUE)
report2 <- last_report()
```

## Cardinality Guards

A join that was one-to-one in development can become many-to-many in production
when upstream data changes. `join_strict()` prevents this by enforcing a
cardinality constraint. If the constraint is violated, the function throws an
error instead of silently producing a Cartesian product.

The workflow has two phases. In development, you use `detect_cardinality()` to
understand the actual relationship:

```{r}
products <- data.frame(
  product_id = c("P1", "P2", "P3"),
  name = c("Widget", "Gadget", "Gizmo"),
  stringsAsFactors = FALSE
)

line_items <- data.frame(
  product_id = c("P1", "P1", "P2", "P3", "P3"),
  order_id = c(101, 102, 103, 104, 105),
  stringsAsFactors = FALSE
)

detect_cardinality(products, line_items, by = "product_id")
```

The result tells you this is a one-to-many relationship: each product appears
once in `products` but can appear multiple times in `line_items`. That is
expected because products are a reference table and line items are transactional. Now
you encode that expectation in production:

```{r}
result <- join_strict(
  products, line_items,
  by = "product_id",
  type = "left",
  expect = "1:m"
)
nrow(result)
```

If someone accidentally loads a `products` table with duplicate product IDs
(turning the relationship into many-to-many), the script fails immediately:

```{r error = TRUE}
products_bad <- data.frame(
  product_id = c("P1", "P1", "P2", "P3"),
  name = c("Widget", "Widget v2", "Gadget", "Gizmo"),
  stringsAsFactors = FALSE
)

join_strict(
  products_bad, line_items,
  by = "product_id",
  type = "left",
  expect = "1:m"
)
```

The four cardinality levels each correspond to a real data pattern:

- **1:1** means a lookup table joined to another lookup table. Each key appears
  exactly once on both sides (country codes to currency codes, employee IDs to
  badge numbers). If either side has duplicates, something went wrong in the
  ETL.

- **1:m** applies when the left table is a reference and the right table has
  transactions or measurements: products to order line items, customers to
  support tickets, stations to hourly weather readings.

- **m:1** is the mirror image: the left table has transactions while the right
  table serves as the lookup. Examples include sales records joined to a region
  table, or survey responses joined to a demographic codebook.

- **m:m** indicates duplicates on both sides. This is almost always a bug. The only
  legitimate use case is generating all pairwise combinations of two attribute
  sets, which is rare enough that requiring an explicit `expect = "m:m"` acts as
  a speed bump.

In practice, `"1:m"` and `"m:1"` cover the vast majority of production joins.
Use `detect_cardinality()` once during development to confirm the relationship,
then hard-code the `expect` value in your production script. The constraint
costs almost nothing at runtime (it checks for duplicates in the key columns
before performing the join) yet it catches a class of upstream data quality
problems that would otherwise propagate silently.

Why not just use `check_cartesian()` instead? The two functions solve different
problems. `check_cartesian()` warns about Cartesian product explosion, the
situation where a key has many duplicates on both sides and the join result
grows quadratically. `join_strict()` enforces a relationship contract. A join
can violate a `"1:1"` constraint without triggering a Cartesian explosion (one
extra duplicate is enough). Conversely, a `"m:m"` join can produce a massive
Cartesian product that `join_strict()` would allow because `"m:m"` is the
loosest constraint. Use `join_strict()` when you know the expected relationship.
Use `check_cartesian()` as an additional guard when the expected relationship
is `"m:m"` and you want to bound the expansion factor.

## Logging and Audit Trails

Production scripts need logs. joinspy provides two mechanisms: manual logging
with `log_report()` and automatic logging with `set_log_file()`.

### Manual logging

`log_report()` writes a single report to a file. The format depends on the
file extension:

```{r}
report <- join_spy(sensors, readings, by = "sensor_id")

# Text format -- human-readable
txt_log <- tempfile(fileext = ".log")
log_report(report, txt_log)
cat(readLines(txt_log), sep = "\n")
unlink(txt_log)
```

```{r}
# JSON format -- machine-readable
json_log <- tempfile(fileext = ".json")
log_report(report, json_log)
cat(readLines(json_log), sep = "\n")
unlink(json_log)
```

Text format is useful when humans read the logs directly: tailing a log file
during a batch run, reviewing output after a nightly job. JSON format is useful
when the logs feed into a monitoring system, a database, or a downstream script
that parses them programmatically. Pick one based on who or what will consume
the log.

You can also save reports as `.rds` files, which preserves the full R object
including all nested lists and data frames. This is heavier than text or JSON
but allows you to reload the report later and inspect it interactively with
`print()`, `summary()`, or direct list access. Use `.rds` when you need the
complete report for post-mortem analysis; use text or JSON when you need a
human-readable or machine-parseable record.

### Automatic logging

For scripts that perform many joins, setting up a log file once at the top is
cleaner than calling `log_report()` after each join. `set_log_file()` enables
automatic logging: every subsequent `*_join_spy()` call appends its report to
the specified file.

```{r}
auto_log <- tempfile(fileext = ".log")
set_log_file(auto_log, format = "text")

# These joins are automatically logged
result1 <- left_join_spy(sensors, readings, by = "sensor_id", .quiet = TRUE)
result2 <- inner_join_spy(sensors, readings, by = "sensor_id", .quiet = TRUE)

# Check what got logged
cat(readLines(auto_log), sep = "\n")

# Clean up
set_log_file(NULL)
unlink(auto_log)
```

The pattern for a production script is: set up the log file at the start, run
all your joins, and disable logging at the end (or let R's session cleanup
handle it). The log accumulates a timestamped record of every join, including
match rates, issue counts, and expected row counts. If something goes wrong
downstream, you can trace back through the log to see which join introduced the
problem.

Automatic logging only triggers from `*_join_spy()` wrappers. If you call
`join_strict()` or bare `merge()` directly, nothing is logged. This is by
design: the wrappers are the "instrumented" path, and `join_strict()` is a
lightweight guard that does not build a full report. If you want both
cardinality enforcement and logging, run `detect_cardinality()` as a separate
check, then use `left_join_spy()` (or whichever wrapper fits) for the actual
join. The complete production pattern at the end of this vignette demonstrates
exactly this approach.

`get_log_file()` returns the current log path (or `NULL` if logging is
disabled), which is handy for conditional logic:

```{r}
# Only log if logging is configured
if (!is.null(get_log_file())) {
  message("Logging is active at: ", get_log_file())
}
```

## Sampling for Large Datasets

Running full diagnostics on a table with ten million rows takes time. The
`sample` parameter in `join_spy()` runs the analysis on a random subset while
the actual join (if you use a `*_join_spy()` wrapper) still operates on the
full data.

```{r}
# Simulate a large dataset
set.seed(42)
big_orders <- data.frame(
  customer_id = sample(paste0("C", sprintf("%04d", 1:5000)), 50000, replace = TRUE),
  amount = round(runif(50000, 10, 500), 2),
  stringsAsFactors = FALSE
)

big_customers <- data.frame(
  customer_id = paste0("C", sprintf("%04d", 1:6000)),
  region = sample(c("North", "South", "East", "West"), 6000, replace = TRUE),
  stringsAsFactors = FALSE
)

# Full analysis
system.time(report_full <- join_spy(big_orders, big_customers, by = "customer_id"))

# Sampled analysis
system.time(report_sampled <- join_spy(big_orders, big_customers,
                                        by = "customer_id", sample = 5000))
```

The sampled report is approximate. Match rates and duplicate counts are
estimated from the sample rather than computed exactly. For most production
monitoring, this is fine; you want to know whether the match rate is roughly
95% or roughly 60%, not whether it is 94.7% or 95.1%. The sample catches
systemic problems (wrong key column, widespread encoding issues, massive
duplicate explosion) with a fraction of the runtime.

The tradeoff: sampling can miss rare issues. If 0.1% of your keys have a
zero-width space, a 5,000-row sample from a 10-million-row table might not
include any of them. For this reason, run full diagnostics periodically (weekly,
or whenever the upstream data source changes) and use sampled diagnostics for
the daily runs. This layered approach gives you speed on most executions and
thoroughness when it matters.

How large should the sample be? There is no universal answer, but a sample of
10,000 to 50,000 rows catches most systemic issues. The goal is not statistical
significance; it is catching problems that affect a meaningful fraction of
keys. If your pipeline joins ten tables, sampling each one at 10,000 rows keeps
the total diagnostic overhead under a second even on modest hardware. The
diagnostics are always cheaper than the join itself, so the sample size rarely
needs to be tuned aggressively.

## A Complete Production Pattern

The sections above introduced each tool individually. Here they are wired
together into a single script that represents a realistic production workflow.
The scenario: a nightly job loads order and customer data, validates keys,
repairs if needed, joins with cardinality enforcement, and logs everything for
audit.

```{r}
# ============================================================
# Nightly order enrichment pipeline
# ============================================================

# --- Setup logging ---
pipeline_log <- tempfile(fileext = ".log")
set_log_file(pipeline_log, format = "text")

# --- Load data (simulated) ---
orders <- data.frame(
  order_id = 1:6,
  customer_id = c("C001", "C002 ", "C003", "C003", "C004", "C005"),
  amount = c(150, 230, 89, 410, 320, 175),
  stringsAsFactors = FALSE
)

customers <- data.frame(
  customer_id = c("C001", "C002", "C003", "C004", "C005", "C006"),
  name = c("Acme Corp", "Globex", "Initech", "Umbrella", "Soylent", "Wonka"),
  tier = c("gold", "silver", "gold", "bronze", "silver", "gold"),
  stringsAsFactors = FALSE
)

# --- Gate 1: key quality assertion ---
keys_ok <- key_check(orders, customers, by = "customer_id", warn = FALSE)

if (!keys_ok) {
  message("Key issues detected -- attempting repair")
  repaired <- join_repair(
    orders, customers,
    by = "customer_id",
    trim_whitespace = TRUE,
    remove_invisible = TRUE
  )
  orders <- repaired$x
  customers <- repaired$y
}

# --- Gate 2: cardinality check ---
card <- detect_cardinality(orders, customers, by = "customer_id")
if (card == "m:m") {
  set_log_file(NULL)
  unlink(pipeline_log)
  stop("Unexpected m:m cardinality in orders-customers join", call. = FALSE)
}

# --- Join (with auto-logging via *_join_spy) ---
enriched <- left_join_spy(orders, customers, by = "customer_id", .quiet = TRUE)

# --- Gate 3: row count sanity check ---
# A left join should never lose rows from the left table
if (nrow(enriched) < nrow(orders)) {
  set_log_file(NULL)
  unlink(pipeline_log)
  stop("Row count decreased after left join -- possible data corruption",
       call. = FALSE)
}

# --- Output ---
message(sprintf("Pipeline complete: %d enriched orders", nrow(enriched)))
head(enriched)

# --- Review the log ---
if (file.exists(pipeline_log)) {
  cat(readLines(pipeline_log), sep = "\n")
}

# --- Cleanup ---
set_log_file(NULL)
unlink(pipeline_log)
```

Each gate in this script serves a different purpose. The `key_check()` gate
catches string-level problems (whitespace, encoding, case) and attempts an
automatic repair. The `detect_cardinality()` gate checks the relationship
between the two tables: orders to customers should be many-to-one, and if the
customer table suddenly has duplicates (a common ETL failure), the script halts
before producing a Cartesian product. The row count check is a final sanity
guard: a left join should preserve all rows from the left table, and if it does
not, something has gone wrong that the other gates did not catch.

The logging runs in the background throughout because `set_log_file()` was
called at the top. Every `*_join_spy()` call appends a timestamped report to
the log file. After the script finishes, the log contains a complete record of
every join: what was checked, what was found, and what the predicted row counts
were. This audit trail pays for itself when debugging a data quality issue days
after the pipeline ran.

The pattern scales. Add more tables, more gates, more joins, and the structure
stays the same. Set up logging once, assert key quality before each join, enforce
cardinality where you know the expected relationship, and check row counts
after. The overhead is minimal: `key_check()` and `detect_cardinality()` are
both O(n) in the number of rows, and `join_strict()` adds only a duplicate
check on top of the join itself. For million-row tables, add `sample = 50000`
to the `join_spy()` calls if the diagnostics become the bottleneck.

The alternative to all of this is trust. Trust that the upstream data is clean,
trust that the ETL did not introduce duplicates, trust that the key columns
still match. Trust works until it does not, and when it breaks, the cost is
always higher than the cost of the checks would have been.
