---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

joinspy diagnoses join keys before you join. It surfaces the string-level problems that kill match rates (whitespace, case, encoding, typos) and offers one-call repair. Works with base R, dplyr, and data.table.

## String Diagnostics

Most failed joins are not caused by missing data. They fail because the keys
*look* identical but are not: a trailing space, a case mismatch, an invisible
Unicode character. These problems are hard to spot by eye and rarely show up in
`str()` or `summary()`. joinspy was built to catch them.

### Whitespace

Trailing and leading whitespace is the single most common reason keys fail to
match. A column exported from Excel or read from a CSV will often contain
`"London "` alongside `"London"`, and R sees those as two distinct values.

```{r}
sales <- data.frame(
  city = c("London", "Paris ", " Berlin", "Tokyo"),
  revenue = c(500, 300, 450, 600),
  stringsAsFactors = FALSE
)

cities <- data.frame(
  city = c("London", "Paris", "Berlin", "Tokyo", "Madrid"),
  country = c("UK", "France", "Germany", "Japan", "Spain"),
  stringsAsFactors = FALSE
)

report <- join_spy(sales, cities, by = "city")
```

Once printed, the report flags the whitespace problems and tells you exactly
which values are affected. The `match_analysis` section shows what the match rate
would be if those issues were fixed; here, `"Paris "` and `" Berlin"` both
fail to match their clean counterparts, dropping the match rate from 100% to
50%.

Whitespace problems compound when keys span multiple columns. Suppose you join
on both city and district:

```{r}
sales2 <- data.frame(
  city = c("London", "Paris ", "Berlin"),
  district = c("West ", "Central", " Mitte"),
  revenue = c(500, 300, 450),
  stringsAsFactors = FALSE
)

districts <- data.frame(
  city = c("London", "Paris", "Berlin"),
  district = c("West", "Central", "Mitte"),
  pop = c(200000, 350000, 180000),
  stringsAsFactors = FALSE
)

report <- join_spy(sales2, districts, by = c("city", "district"))
```

Now London fails on the district column and Paris fails on the city column.
Each column is checked independently; a row needs all key columns to match, so
a single trailing space in any one of them is enough to break the join.

For a quick pass/fail gate in scripts, use `key_check()`:

```{r}
key_check(sales, cities, by = "city")
```

`key_check()` returns `FALSE` here because of the whitespace. When
`warn = TRUE` (the default), it also prints which issues it found. This makes
`key_check()` useful as an assertion in automated pipelines: wrap it in
`stopifnot()` and the script halts before a broken join silently drops rows.
Unlike `join_spy()`, which returns a full report object, `key_check()` is
deliberately minimal (a single logical value) so it slots cleanly into
conditional logic.

### Case Mismatches

Databases and spreadsheets handle case differently. A CRM might store `"ACME"` while the billing system stores `"Acme"`, and R treats these as distinct strings.

```{r}
invoices <- data.frame(
  company = c("ACME", "globex", "Initech", "UMBRELLA"),
  amount = c(1200, 800, 950, 1500),
  stringsAsFactors = FALSE
)

vendors <- data.frame(
  company = c("Acme", "Globex", "Initech", "Umbrella"),
  sector = c("Manufacturing", "Logistics", "Software", "Biotech"),
  stringsAsFactors = FALSE
)

report <- join_spy(invoices, vendors, by = "company")
```

Each case-mismatched pair is listed in the output. Only `"Initech"` matches exactly;
the other three would be lost in a standard join. Case conflicts are particularly common
when merging data from systems with different conventions: CRMs that uppercase
everything, ERPs that title-case, web forms that accept freeform input. The
`match_analysis` section of the report shows the potential match rate both with
and without case sensitivity, so you can gauge whether standardizing case would
recover most of the lost matches.

### Encoding and Invisible Characters

Encoding issues are the nastiest. Nothing visible to debug. Two strings render identically in the console but differ at the byte level: a non-breaking space (`\u00A0`) versus a regular space, a Latin-1 accented character versus its UTF-8 equivalent.

```{r}
# Simulate invisible character contamination
raw_ids <- data.frame(
  product_id = c("SKU-001", "SKU-002", paste0("SKU-003", "\u200B")),
  batch = c("A", "B", "C"),
  stringsAsFactors = FALSE
)

clean_ids <- data.frame(
  product_id = c("SKU-001", "SKU-002", "SKU-003"),
  warehouse = c("East", "West", "North"),
  stringsAsFactors = FALSE
)

report <- join_spy(raw_ids, clean_ids, by = "product_id")
```

The zero-width space appended to `"SKU-003"` is invisible when printed but
prevents the join from matching. joinspy flags this under encoding issues. Other
common culprits include non-breaking spaces (`\u00A0`), byte-order marks, and
right-to-left marks that creep in from PDF copy-paste or web scraping. Standard
R string functions (`trimws()`, `nchar()`) will not catch these because they
operate on visible characters only. By identifying the specific Unicode
code points involved, joinspy saves considerable debugging time compared to
staring at two strings that look identical in the console.

### Combining Multiple Issues

Real data rarely has just one problem. A single key column might have trailing whitespace on some rows, case mismatches on others, invisible characters on a third set. `join_spy()` checks for all of these in one call and reports them together, grouped by type.

```{r}
employees <- data.frame(
  dept = c("Sales ", "ENGINEERING", "marketing", paste0("HR", "\u00A0")),
  name = c("Alice", "Bob", "Carol", "David"),
  stringsAsFactors = FALSE
)

departments <- data.frame(
  dept = c("Sales", "Engineering", "Marketing", "HR"),
  budget = c(50000, 80000, 45000, 30000),
  stringsAsFactors = FALSE
)

report <- join_spy(employees, departments, by = "dept")
```

Zero matches out of four rows. Grouped by problem type (whitespace, case, encoding), the output lets you
decide how to address each one. This
matters because the fix differs: whitespace is safe to trim automatically, case
standardization requires choosing a direction (upper, lower, or title), and
encoding issues sometimes indicate deeper data pipeline problems that trimming
alone will not solve. The grouped output lets you prioritize; often fixing
whitespace and case recovers 90% of lost matches, and the remaining encoding
problems can be handled case by case.

### Duplicate Keys

Duplicates do not prevent matches, but they multiply rows. If `"London"`
appears three times in the left table and twice in the right, you get six rows
for London after an inner join, often unintentionally.

`key_duplicates()` returns the offending rows along with a count:

```{r}
transactions <- data.frame(
  store_id = c("S1", "S1", "S2", "S3", "S3", "S3"),
  day = c("Mon", "Tue", "Mon", "Mon", "Tue", "Wed"),
  amount = c(100, 200, 150, 300, 250, 400),
  stringsAsFactors = FALSE
)

key_duplicates(transactions, by = "store_id")
```

Use `keep = "first"` or `keep = "last"` to return only the first or last
occurrence per key. The default `keep = "all"` returns every row that
participates in a duplicate group. The `.n_duplicates` column in the output
tells you the group size, which directly maps to the row multiplication factor:
a key with `.n_duplicates = 3` in the left table and `.n_duplicates = 2` in the
right will produce 6 rows for that key in an inner join.

## Auto-Repair

Once you know what is wrong, joinspy can fix it. `join_repair()` handles
whitespace trimming, case standardization, invisible character removal, and
empty-to-NA conversion in a single call.

### Dry Run

Always preview repairs before applying them. The `dry_run = TRUE` option shows
what would change without modifying the data:

```{r}
messy_left <- data.frame(
  id = c(" A-101", "B-202 ", "c-303", paste0("D-404", "\u200B")),
  score = c(88, 92, 76, 95),
  stringsAsFactors = FALSE
)

messy_right <- data.frame(
  id = c("A-101", "B-202", "C-303", "D-404"),
  label = c("Alpha", "Beta", "Gamma", "Delta"),
  stringsAsFactors = FALSE
)

join_repair(messy_left, messy_right, by = "id", dry_run = TRUE)
```

The dry run tells you how many values will change and in which direction,
without touching the original data frames. Review this output carefully: it
shows which values will be modified and what they will become. Review
carefully for case standardization, where uppercasing a key like
`"c-303"` to `"C-303"` is correct, but you want to verify that both tables
agree on the target format before committing.

### Applying Repairs

Drop `dry_run` (or set it to `FALSE`) to get back the repaired data frames.
When you pass both `x` and `y`, the return value is a list with both:

```{r}
repaired <- join_repair(
  messy_left, messy_right,
  by = "id",
  trim_whitespace = TRUE,
  standardize_case = "upper",
  remove_invisible = TRUE
)

repaired$x$id
repaired$y$id
```

Both sides now have clean, uppercase, whitespace-free keys. You can pass the
repaired tables straight into a join.

To confirm the repairs worked, run `key_check()` on the repaired tables:

```{r}
key_check(repaired$x, repaired$y, by = "id")
```

This should return `TRUE` with no warnings. Running `key_check()` after repair
is good practice because it catches edge cases where trimming or case
standardization was not enough (for instance, if two distinct keys collapse into
the same value after uppercasing).

One important caveat: `join_repair()` fixes mechanical string problems:
whitespace, case, invisible characters, empty strings. It does not fix typos,
semantic mismatches, or fundamentally different key formats. If one table uses
`"USA"` and the other uses `"United States"`, or one stores product codes as
`"P-001"` while the other uses `"001"`, repair will not bridge that gap. Those
require manual recoding or a lookup table.

### Repair Suggestions from a Report

If you already have a `JoinReport` from `join_spy()`, call `suggest_repairs()`
to get ready-to-run R code that fixes the detected issues:

```{r}
report <- join_spy(messy_left, messy_right, by = "id")
suggest_repairs(report)
```

This prints code snippets you can copy into your script. It is deliberately not
automatic: you should review the suggestions before running them. The generated
code calls `join_repair()` with the appropriate flags enabled based on what the
report detected, so you do not have to remember which issues were present.

## Row Count Predictions

Every `JoinReport` includes an `expected_rows` component that predicts how many
rows each join type will produce, calculated from the key overlap and
duplicate structure, before any join is executed.

```{r}
orders <- data.frame(
  product_id = c("P1", "P1", "P2", "P3", "P4"),
  quantity = c(10, 5, 20, 15, 8),
  stringsAsFactors = FALSE
)

products <- data.frame(
  product_id = c("P1", "P2", "P3", "P5"),
  name = c("Widget", "Gadget", "Gizmo", "Doohickey"),
  stringsAsFactors = FALSE
)

report <- join_spy(orders, products, by = "product_id")
report$expected_rows
```

The predictions help you anticipate whether a join will shrink, grow, or
explode your data. A left join here should retain all five order rows (with
`NA` for the unmatched `"P4"`), while an inner join drops it.

How does joinspy compute these numbers? For each join type, it counts the
overlapping keys between the two tables and multiplies by the number of
duplicates on each side. A key that appears twice in `x` and three times in `y`
contributes 2 * 3 = 6 rows to an inner join. Non-matching keys add rows for
left, right, and full joins according to the usual rules. The calculation is
exact when all keys are non-NA and no string issues remain; however, NA keys
deserve caution. R's `merge()` does not match NAs by default (and neither does
dplyr), so if either table contains NA keys, the predicted row count may
overestimate the actual result. The report warns you when NAs are present in the
key columns so you can account for this.

`summary()` returns the same information as a compact data frame:

```{r}
summary(report)
```

## Post-Join Diagnostics

Sometimes the join has already happened and you need to understand the result.

### Explaining Row Count Changes

`join_explain()` takes the result, the two input tables, and the join column,
then tells you why the row count changed:

```{r}
tickets <- data.frame(
  event_id = c(1, 2, 2, 3),
  seat = c("A1", "B2", "B3", "C1"),
  stringsAsFactors = FALSE
)

events <- data.frame(
  event_id = c(1, 2, 4),
  name = c("Concert", "Play", "Opera"),
  stringsAsFactors = FALSE
)

result <- merge(tickets, events, by = "event_id")
join_explain(result, tickets, events, by = "event_id", type = "inner")
```

The output breaks down the row count into contributions from matched keys,
unmatched keys, and duplicate-driven expansion. Specifically, you get: the
number of keys that matched in both tables; how many rows each matched key
contributed (accounting for duplicates); which keys from each table had no
match; and the net effect on row count compared to the left table. This
decomposition makes it straightforward to answer questions like "why did my
table grow by 200 rows?", and usually the answer is a handful of keys with
unexpected duplicates on one side. If you omit `type`, joinspy infers the join
type from the row counts (though providing it explicitly is more reliable).

### Before/After Comparison

`join_diff()` compares a data frame before and after a join to show which rows
were added, dropped, or preserved:

```{r}
before <- data.frame(
  id = 1:4,
  value = c("a", "b", "c", "d"),
  stringsAsFactors = FALSE
)

lookup <- data.frame(
  id = c(2, 3, 4, 5),
  extra = c("X", "Y", "Z", "W"),
  stringsAsFactors = FALSE
)

after <- merge(before, lookup, by = "id", all = TRUE)
join_diff(before, after, by = "id")
```

Row 1 was dropped (no match in the right table for a non-full join is
irrelevant here since we used `all = TRUE`; but row 5 was added from the right
table). This kind of accounting is tedious to do by hand on large data.

Here is a second example that highlights row addition from duplicates rather
than new keys:

```{r}
before2 <- data.frame(
  id = c(1, 2, 3),
  value = c("a", "b", "c"),
  stringsAsFactors = FALSE
)

dup_lookup <- data.frame(
  id = c(1, 1, 2, 3),
  tag = c("X", "Y", "Z", "W"),
  stringsAsFactors = FALSE
)

after2 <- merge(before2, dup_lookup, by = "id")
join_diff(before2, after2, by = "id")
```

Here the row count grew from three to four because `id = 1` matched two rows
in the lookup. No new keys were introduced; the growth came entirely from
duplicate-driven expansion. `join_diff()` distinguishes these two sources of
row addition (new keys vs. duplicate expansion), which is the key information
you need when debugging an unexpectedly large result.

## Safe Join Wrappers

The `*_join_spy()` family wraps standard joins with automatic diagnostics. They
print a report before joining and return the joined data frame:

```{r}
patients <- data.frame(
  patient_id = c("P01", "P02", "P03"),
  age = c(34, 28, 45),
  stringsAsFactors = FALSE
)

labs <- data.frame(
  patient_id = c("P01", "P02", "P04"),
  result = c(5.2, 3.8, 6.1),
  stringsAsFactors = FALSE
)

result <- left_join_spy(patients, labs, by = "patient_id")
head(result)
```

All four wrappers (`left_join_spy`, `right_join_spy`, `inner_join_spy`,
`full_join_spy`) share the same interface.

When should you use a wrapper instead of calling `join_spy()` manually? The
wrappers are convenient for exploratory work and interactive sessions where you
want diagnostics printed as a side effect without breaking your pipeline. In
production scripts, you may prefer calling `join_spy()` first, inspecting the
report programmatically (checking `report$issues` or `report$expected_rows`),
and only proceeding with the join if the diagnostics pass. The wrappers also
accept a `backend` parameter (`"base"`, `"dplyr"`, or `"data.table"`) that
overrides the automatic backend detection; this is useful when you want to force
a specific join engine regardless of the input class.

### Quiet Mode and Deferred Reports

In pipelines, printed output is distracting. Use `.quiet = TRUE` to suppress
the report, then retrieve it afterward with `last_report()`:

```{r}
result <- inner_join_spy(patients, labs, by = "patient_id", .quiet = TRUE)

# Later, inspect what happened
rpt <- last_report()
rpt$match_analysis$match_rate
```

`last_report()` always returns the report from the most recent `*_join_spy()`
call, regardless of which wrapper was used. The report object is stored in an
internal environment and persists for the duration of the R session, so you can
inspect it at any point after the join -- even after several more lines of
pipeline code have executed.

## Cardinality Enforcement

`join_strict()` performs a join that throws an error if the key relationship
violates your expectation. This is useful as a guard in production scripts
where a surprise many-to-many relationship would corrupt downstream results.

```{r}
sensors <- data.frame(
  sensor_id = c("T1", "T2", "T3"),
  location = c("Roof", "Basement", "Lobby"),
  stringsAsFactors = FALSE
)

readings <- data.frame(
  sensor_id = c("T1", "T2", "T3"),
  value = c(22.1, 18.5, 21.0),
  stringsAsFactors = FALSE
)

# Succeeds: one reading per sensor
join_strict(sensors, readings, by = "sensor_id", expect = "1:1")
```

```{r error = TRUE}
# Fails: T1 has two readings, violating 1:1
readings_dup <- data.frame(
  sensor_id = c("T1", "T1", "T2", "T3"),
  value = c(22.1, 23.4, 18.5, 21.0),
  stringsAsFactors = FALSE
)

join_strict(sensors, readings_dup, by = "sensor_id", expect = "1:1")
```

The `expect` argument accepts four cardinality levels, each appropriate for
different scenarios:

- `"1:1"`: both sides have unique keys. Use this when merging two reference
  tables (e.g., a country table with a currency table, both keyed on country
  code). Any duplicate on either side is an error.
- `"1:m"`: the left table has unique keys, the right may repeat. This is the
  classic lookup-table join: a product catalogue (left) joined to a
  transactions table (right) where each product appears in many transactions.
- `"m:1"`: the mirror of `"1:m"`. The right table is the lookup; the left
  table has repeats.
- `"m:m"`: duplicates are allowed on both sides. This disables the constraint
  entirely and is rarely what you want; it exists as an explicit opt-in for the
  rare cases where a Cartesian product per key is intentional (e.g., generating
  all pairings between two sets of attributes).

To detect the actual cardinality without enforcing anything:

```{r}
detect_cardinality(sensors, readings_dup, by = "sensor_id")
```

This returns a string like `"1:m"` that you can use programmatically. You might
use this in a script to select the right `expect` value for `join_strict()`
dynamically, or to log the cardinality of a join for monitoring purposes.

## Advanced Features

### Cartesian Product Detection

When both tables have duplicates on the same key, a join produces the Cartesian
product of those duplicates. Two keys with 100 rows each? 10,000 rows for that key alone. `check_cartesian()` warns you before this happens:

```{r}
left <- data.frame(
  group = c("A", "A", "A", "B"),
  x = 1:4,
  stringsAsFactors = FALSE
)

right <- data.frame(
  group = c("A", "A", "B", "B"),
  y = 5:8,
  stringsAsFactors = FALSE
)

check_cartesian(left, right, by = "group")
```

The `threshold` argument (default 10) controls how large the expansion factor
must be before a warning is triggered. Set it lower for interactive work where
you want early warnings; set it higher in batch scripts where moderate expansion
is expected and you only want to catch truly pathological cases.

### Multi-Table Join Chains

`analyze_join_chain()` runs diagnostics across a sequence of joins. You provide
a named list of tables and a list of join specifications:

```{r}
orders <- data.frame(
  order_id = 1:4,
  customer_id = c(1, 2, 2, 3),
  stringsAsFactors = FALSE
)

customers <- data.frame(
  customer_id = 1:3,
  region_id = c(1, 1, 2),
  stringsAsFactors = FALSE
)

regions <- data.frame(
  region_id = 1:2,
  name = c("North", "South"),
  stringsAsFactors = FALSE
)

analyze_join_chain(
  tables = list(orders = orders, customers = customers, regions = regions),
  joins = list(
    list(left = "orders", right = "customers", by = "customer_id"),
    list(left = "result", right = "regions", by = "region_id")
  )
)
```

Each step prints a diagnostic report. The special name `"result"` in the
second step refers to the output of the previous join, so you can chain as many
tables as needed. This placeholder is resolved internally: after the first join
produces an intermediate table, `"result"` is swapped in as the left (or right)
table for the next step. This means key diagnostics at step two operate on the
actual intermediate data, not on the original input, which is important because
the first join may have introduced NAs or changed the key distribution.

### Backend Support

All join wrappers and `join_strict()` accept a `backend` argument. By default
(`backend = NULL`), joinspy detects the backend from the input class: tibbles
use dplyr, data.tables use data.table, and plain data frames use base R. You
can override this:

```{r eval = FALSE}
# Force dplyr backend even for plain data frames
left_join_spy(x, y, by = "id", backend = "dplyr")

# Force base R for tibbles
left_join_spy(x, y, by = "id", backend = "base")
```

The three backends are `"base"`, `"dplyr"`, and `"data.table"`. The diagnostic
layer is backend-agnostic; all string analysis and key checking happens before
the join engine is invoked, so the reports are identical regardless of which
backend performs the actual merge.

## Quick Reference

| Function | Purpose | Returns |
|---|---|---|
| `join_spy()` | Full pre-join diagnostic | `JoinReport` (print/summary/plot) |
| `key_check()` | Quick pass/fail | Logical |
| `key_duplicates()` | Find duplicate keys | Data frame with `.n_duplicates` |
| `join_repair()` | Fix whitespace, case, encoding | Repaired data frame(s) |
| `suggest_repairs()` | Print repair code snippets | Printed output (invisible) |
| `left_join_spy()` | Left join + diagnostics | Joined data frame |
| `right_join_spy()` | Right join + diagnostics | Joined data frame |
| `inner_join_spy()` | Inner join + diagnostics | Joined data frame |
| `full_join_spy()` | Full join + diagnostics | Joined data frame |
| `last_report()` | Retrieve last quiet report | `JoinReport` or `NULL` |
| `join_strict()` | Cardinality-enforcing join | Joined data frame or error |
| `join_explain()` | Explain row count change | Printed summary |
| `join_diff()` | Before/after comparison | Printed summary |
| `detect_cardinality()` | Detect key relationship | Character (`"1:1"`, etc.) |
| `check_cartesian()` | Warn about row explosion | List with explosion analysis |
| `analyze_join_chain()` | Multi-step join diagnostics | Printed chain analysis |
| `log_report()` | Write report to file | File output (invisible) |
| `set_log_file()` | Enable auto-logging | Sets global log path |

## Further Reading

- `vignette("common-issues")` for a catalogue of join problems and solutions

- Function help pages: `?join_spy`, `?join_repair`, `?join_strict`
