---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

joinspy diagnoses join keys before you join. It surfaces the string-level problems that kill match rates (whitespace, case, encoding, typos) and offers one-call repair. Works with base R, dplyr, and data.table.

## String Diagnostics

The keys *look* identical but differ at the byte level: a trailing space, a case mismatch, an invisible Unicode character. These differences don't show up in `str()` or `summary()` output, but they determine whether a join matches.

### Whitespace

Trailing and leading whitespace is the most common reason keys fail to match. A column exported from Excel might contain `"London "` alongside `"London"`, and R treats those as distinct values.

```{r}
sales <- data.frame(
  city = c("London", "Paris ", " Berlin", "Tokyo"),
  revenue = c(500, 300, 450, 600),
  stringsAsFactors = FALSE
)

cities <- data.frame(
  city = c("London", "Paris", "Berlin", "Tokyo", "Madrid"),
  country = c("UK", "France", "Germany", "Japan", "Spain"),
  stringsAsFactors = FALSE
)

report <- join_spy(sales, cities, by = "city")
```

The report flags the whitespace problems and shows which values are affected. `"Paris "` and `" Berlin"` both fail to match, dropping the match rate from 100% to 50%.

Whitespace problems compound with multi-column keys:

```{r}
sales2 <- data.frame(
  city = c("London", "Paris ", "Berlin"),
  district = c("West ", "Central", " Mitte"),
  revenue = c(500, 300, 450),
  stringsAsFactors = FALSE
)

districts <- data.frame(
  city = c("London", "Paris", "Berlin"),
  district = c("West", "Central", "Mitte"),
  pop = c(200000, 350000, 180000),
  stringsAsFactors = FALSE
)

report <- join_spy(sales2, districts, by = c("city", "district"))
```

London fails on the district column, Paris on the city column. A row needs all key columns to match, so a single trailing space in any column breaks the join.

For a quick pass/fail gate, we can use `key_check()`:

```{r}
key_check(sales, cities, by = "city")
```

`key_check()` returns `FALSE` here because of the whitespace, and with `warn = TRUE` (the default) it prints which issues it found. It returns a single logical, so it slots into `stopifnot()` or conditional logic.

### Case Mismatches

A CRM might store `"ACME"` while the billing system stores `"Acme"`, and R treats these as distinct strings.

```{r}
invoices <- data.frame(
  company = c("ACME", "globex", "Initech", "UMBRELLA"),
  amount = c(1200, 800, 950, 1500),
  stringsAsFactors = FALSE
)

vendors <- data.frame(
  company = c("Acme", "Globex", "Initech", "Umbrella"),
  sector = c("Manufacturing", "Logistics", "Software", "Biotech"),
  stringsAsFactors = FALSE
)

report <- join_spy(invoices, vendors, by = "company")
```

Only `"Initech"` matches exactly; the other three would be lost in a standard join. The `match_analysis` section shows the potential match rate both with and without case sensitivity.

### Encoding and Invisible Characters

Two strings can render identically in the console but differ at the byte level: a non-breaking space (`\u00A0`) versus a regular space, a zero-width space from PDF copy-paste.

```{r}
# Simulate invisible character contamination
raw_ids <- data.frame(
  product_id = c("SKU-001", "SKU-002", paste0("SKU-003", "\u200B")),
  batch = c("A", "B", "C"),
  stringsAsFactors = FALSE
)

clean_ids <- data.frame(
  product_id = c("SKU-001", "SKU-002", "SKU-003"),
  warehouse = c("East", "West", "North"),
  stringsAsFactors = FALSE
)

report <- join_spy(raw_ids, clean_ids, by = "product_id")
```

The zero-width space appended to `"SKU-003"` is invisible when printed but prevents the match. joinspy flags this under encoding issues and identifies the specific Unicode code points involved.

### Combining Multiple Issues

Real data rarely has just one problem. `join_spy()` checks for whitespace, case, and encoding issues in one call and reports them together, grouped by type.

```{r}
employees <- data.frame(
  dept = c("Sales ", "ENGINEERING", "marketing", paste0("HR", "\u00A0")),
  name = c("Alice", "Bob", "Carol", "David"),
  stringsAsFactors = FALSE
)

departments <- data.frame(
  dept = c("Sales", "Engineering", "Marketing", "HR"),
  budget = c(50000, 80000, 45000, 30000),
  stringsAsFactors = FALSE
)

report <- join_spy(employees, departments, by = "dept")
```

Zero matches out of four rows, with the problems grouped by type. The fix differs per category: whitespace is safe to trim automatically, case standardization requires choosing a direction, and encoding issues may indicate deeper pipeline problems.

### Duplicate Keys

Duplicates don't prevent matches, but they multiply rows. If `"London"` appears three times in the left table and twice in the right, we get six rows for London after an inner join.

`key_duplicates()` returns the offending rows along with a count:

```{r}
transactions <- data.frame(
  store_id = c("S1", "S1", "S2", "S3", "S3", "S3"),
  day = c("Mon", "Tue", "Mon", "Mon", "Tue", "Wed"),
  amount = c(100, 200, 150, 300, 250, 400),
  stringsAsFactors = FALSE
)

key_duplicates(transactions, by = "store_id")
```

The `keep` argument accepts `"first"`, `"last"`, or the default `"all"`. The `.n_duplicates` column maps directly to the row multiplication factor in a join.

## Auto-Repair

`join_repair()` handles whitespace trimming, case standardization, invisible character removal, and empty-to-NA conversion in a single call.

### Dry Run

We can preview what `join_repair()` would change with `dry_run = TRUE`:

```{r}
messy_left <- data.frame(
  id = c(" A-101", "B-202 ", "c-303", paste0("D-404", "\u200B")),
  score = c(88, 92, 76, 95),
  stringsAsFactors = FALSE
)

messy_right <- data.frame(
  id = c("A-101", "B-202", "C-303", "D-404"),
  label = c("Alpha", "Beta", "Gamma", "Delta"),
  stringsAsFactors = FALSE
)

join_repair(messy_left, messy_right, by = "id", dry_run = TRUE)
```

The dry run shows how many values will change and what they become, without touching the original data.

### Applying Repairs

Drop `dry_run` (or set it to `FALSE`) to get back the repaired data frames. When we pass both `x` and `y`, the return value is a list with both:

```{r}
repaired <- join_repair(
  messy_left, messy_right,
  by = "id",
  trim_whitespace = TRUE,
  standardize_case = "upper",
  remove_invisible = TRUE
)

repaired$x$id
repaired$y$id
```

Both sides now have clean, uppercase, whitespace-free keys. We can confirm with `key_check()`:

```{r}
key_check(repaired$x, repaired$y, by = "id")
```

`join_repair()` fixes mechanical string problems: whitespace, case, invisible characters, empty strings. It does not fix typos, semantic mismatches, or different key formats (`"USA"` vs. `"United States"`).

### Repair Suggestions from a Report

If we already have a `JoinReport` from `join_spy()`, `suggest_repairs()` generates ready-to-run code that fixes the detected issues:

```{r}
report <- join_spy(messy_left, messy_right, by = "id")
suggest_repairs(report)
```

The generated code calls `join_repair()` with the appropriate flags based on what the report detected.

## Row Count Predictions

Every `JoinReport` includes an `expected_rows` component that predicts how many rows each join type will produce, calculated from key overlap and duplicate structure.

```{r}
orders <- data.frame(
  product_id = c("P1", "P1", "P2", "P3", "P4"),
  quantity = c(10, 5, 20, 15, 8),
  stringsAsFactors = FALSE
)

products <- data.frame(
  product_id = c("P1", "P2", "P3", "P5"),
  name = c("Widget", "Gadget", "Gizmo", "Doohickey"),
  stringsAsFactors = FALSE
)

report <- join_spy(orders, products, by = "product_id")
report$expected_rows
```

A left join here retains all five order rows (with `NA` for the unmatched `"P4"`), while an inner join drops it.

For each join type, the calculation counts overlapping keys and multiplies by the number of duplicates on each side. A key appearing twice in `x` and three times in `y` contributes 2 * 3 = 6 rows to an inner join. The calculation is exact when all keys are non-NA and no string issues remain. NA keys deserve caution: R's `merge()` and dplyr both skip NAs by default, so the predicted count may overestimate. The report warns when NAs are present.

`summary()` returns the same information as a compact data frame:

```{r}
summary(report)
```

## Post-Join Diagnostics

Sometimes the join has already happened and we need to understand the result.

### Explaining Row Count Changes

`join_explain()` takes the result, the two input tables, and the join column, then explains why the row count changed:

```{r}
tickets <- data.frame(
  event_id = c(1, 2, 2, 3),
  seat = c("A1", "B2", "B3", "C1"),
  stringsAsFactors = FALSE
)

events <- data.frame(
  event_id = c(1, 2, 4),
  name = c("Concert", "Play", "Opera"),
  stringsAsFactors = FALSE
)

result <- merge(tickets, events, by = "event_id")
join_explain(result, tickets, events, by = "event_id", type = "inner")
```

The output decomposes the row count into contributions from matched keys, unmatched keys, and duplicate-driven expansion. Usually when a table grows unexpectedly, the answer is a handful of keys with duplicates on one side.

### Before/After Comparison

`join_diff()` compares a data frame before and after a join to show which rows were added, dropped, or preserved:

```{r}
before <- data.frame(
  id = 1:4,
  value = c("a", "b", "c", "d"),
  stringsAsFactors = FALSE
)

lookup <- data.frame(
  id = c(2, 3, 4, 5),
  extra = c("X", "Y", "Z", "W"),
  stringsAsFactors = FALSE
)

after <- merge(before, lookup, by = "id", all = TRUE)
join_diff(before, after, by = "id")
```

Row 1 was dropped (no match in the right table for a non-full join is irrelevant here since we used `all = TRUE`; but row 5 was added from the right table).

Here is a second example that highlights row addition from duplicates rather than new keys:

```{r}
before2 <- data.frame(
  id = c(1, 2, 3),
  value = c("a", "b", "c"),
  stringsAsFactors = FALSE
)

dup_lookup <- data.frame(
  id = c(1, 1, 2, 3),
  tag = c("X", "Y", "Z", "W"),
  stringsAsFactors = FALSE
)

after2 <- merge(before2, dup_lookup, by = "id")
join_diff(before2, after2, by = "id")
```

The row count grew from three to four because `id = 1` matched two rows in the lookup. No new keys were introduced; the growth came entirely from duplicate expansion. `join_diff()` distinguishes these two sources of row addition (new keys vs. duplicate expansion).

## Safe Join Wrappers

The `*_join_spy()` family wraps standard joins with automatic diagnostics. They print a report before joining and return the joined data frame:

```{r}
patients <- data.frame(
  patient_id = c("P01", "P02", "P03"),
  age = c(34, 28, 45),
  stringsAsFactors = FALSE
)

labs <- data.frame(
  patient_id = c("P01", "P02", "P04"),
  result = c(5.2, 3.8, 6.1),
  stringsAsFactors = FALSE
)

result <- left_join_spy(patients, labs, by = "patient_id")
head(result)
```

All four wrappers (`left_join_spy`, `right_join_spy`, `inner_join_spy`, `full_join_spy`) share the same interface. The wrappers are convenient for exploratory work; in production, we might prefer calling `join_spy()` first and inspecting the report programmatically before proceeding. The wrappers also accept a `backend` parameter (`"base"`, `"dplyr"`, or `"data.table"`) that overrides automatic backend detection.

### Quiet Mode and Deferred Reports

In pipelines, printed output can be distracting. With `.quiet = TRUE`, the report is suppressed and can be retrieved afterward via `last_report()`:

```{r}
result <- inner_join_spy(patients, labs, by = "patient_id", .quiet = TRUE)

# Later, inspect what happened
rpt <- last_report()
rpt$match_analysis$match_rate
```

`last_report()` returns the report from the most recent `*_join_spy()` call and persists for the duration of the R session.

## Cardinality Enforcement

`join_strict()` performs a join that errors if the key relationship violates a stated expectation. This is a guard against unexpected many-to-many joins that multiply rows.

```{r}
sensors <- data.frame(
  sensor_id = c("T1", "T2", "T3"),
  location = c("Roof", "Basement", "Lobby"),
  stringsAsFactors = FALSE
)

readings <- data.frame(
  sensor_id = c("T1", "T2", "T3"),
  value = c(22.1, 18.5, 21.0),
  stringsAsFactors = FALSE
)

# Succeeds: one reading per sensor
join_strict(sensors, readings, by = "sensor_id", expect = "1:1")
```

```{r error = TRUE}
# Fails: T1 has two readings, violating 1:1
readings_dup <- data.frame(
  sensor_id = c("T1", "T1", "T2", "T3"),
  value = c(22.1, 23.4, 18.5, 21.0),
  stringsAsFactors = FALSE
)

join_strict(sensors, readings_dup, by = "sensor_id", expect = "1:1")
```

The `expect` argument accepts four cardinality levels:

- `"1:1"`: both sides have unique keys. Any duplicate on either side is an error.
- `"1:m"`: the left table has unique keys, the right may repeat. The classic lookup-table join.
- `"m:1"`: the mirror of `"1:m"`. The right table is the lookup.
- `"m:m"`: duplicates allowed on both sides. Rarely what we want; exists as an explicit opt-in for intentional Cartesian products per key.

To detect the actual cardinality without enforcing anything:

```{r}
detect_cardinality(sensors, readings_dup, by = "sensor_id")
```

This returns a string like `"1:m"` that we can use programmatically, e.g. to select the right `expect` value or to log the cardinality for monitoring.

## Advanced Features

### Cartesian Product Detection

When both tables have duplicates on the same key, the join produces the Cartesian product of those duplicates. Two keys with 100 rows each produce 10,000 rows for that key alone. `check_cartesian()` warns before this happens:

```{r}
left <- data.frame(
  group = c("A", "A", "A", "B"),
  x = 1:4,
  stringsAsFactors = FALSE
)

right <- data.frame(
  group = c("A", "A", "B", "B"),
  y = 5:8,
  stringsAsFactors = FALSE
)

check_cartesian(left, right, by = "group")
```

The `threshold` argument (default 10) controls how large the expansion factor must be before a warning fires.

### Multi-Table Join Chains

`analyze_join_chain()` runs diagnostics across a sequence of joins. We provide a named list of tables and a list of join specifications:

```{r}
orders <- data.frame(
  order_id = 1:4,
  customer_id = c(1, 2, 2, 3),
  stringsAsFactors = FALSE
)

customers <- data.frame(
  customer_id = 1:3,
  region_id = c(1, 1, 2),
  stringsAsFactors = FALSE
)

regions <- data.frame(
  region_id = 1:2,
  name = c("North", "South"),
  stringsAsFactors = FALSE
)

analyze_join_chain(
  tables = list(orders = orders, customers = customers, regions = regions),
  joins = list(
    list(left = "orders", right = "customers", by = "customer_id"),
    list(left = "result", right = "regions", by = "region_id")
  )
)
```

Each step prints a diagnostic report. The special name `"result"` refers to the output of the previous join, so we can chain as many tables as needed. Because `"result"` resolves to the actual intermediate data, key diagnostics at step two account for any NAs or key distribution changes introduced by the first join.

### Backend Support

All join wrappers and `join_strict()` accept a `backend` argument. By default (`backend = NULL`), joinspy detects the backend from the input class: tibbles use dplyr, data.tables use data.table, plain data frames use base R.

```{r eval = FALSE}
# Force dplyr backend even for plain data frames
left_join_spy(x, y, by = "id", backend = "dplyr")

# Force base R for tibbles
left_join_spy(x, y, by = "id", backend = "base")
```

The diagnostic layer is backend-agnostic; all string analysis and key checking happens before the join engine is invoked, so reports are identical regardless of backend.

## Quick Reference

| Function | Purpose | Returns |
|---|---|---|
| `join_spy()` | Full pre-join diagnostic | `JoinReport` (print/summary/plot) |
| `key_check()` | Quick pass/fail | Logical |
| `key_duplicates()` | Find duplicate keys | Data frame with `.n_duplicates` |
| `join_repair()` | Fix whitespace, case, encoding | Repaired data frame(s) |
| `suggest_repairs()` | Print repair code snippets | Printed output (invisible) |
| `left_join_spy()` | Left join + diagnostics | Joined data frame |
| `right_join_spy()` | Right join + diagnostics | Joined data frame |
| `inner_join_spy()` | Inner join + diagnostics | Joined data frame |
| `full_join_spy()` | Full join + diagnostics | Joined data frame |
| `last_report()` | Retrieve last quiet report | `JoinReport` or `NULL` |
| `join_strict()` | Cardinality-enforcing join | Joined data frame or error |
| `join_explain()` | Explain row count change | Printed summary |
| `join_diff()` | Before/after comparison | Printed summary |
| `detect_cardinality()` | Detect key relationship | Character (`"1:1"`, etc.) |
| `check_cartesian()` | Warn about row explosion | List with explosion analysis |
| `analyze_join_chain()` | Multi-step join diagnostics | Printed chain analysis |
| `log_report()` | Write report to file | File output (invisible) |
| `set_log_file()` | Enable auto-logging | Sets global log path |

## Further Reading

- `vignette("common-issues")` for a catalogue of join problems and solutions

- Function help pages: `?join_spy`, `?join_repair`, `?join_strict`
