---
title: "Why Your Keys Don't Match"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why Your Keys Don't Match}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(joinspy)
```

You write a join. It runs without error. You check the row count and something
is off: rows vanished, or they multiplied. The columns look fine. The key
values look identical when you print them. You stare at `str()` output for ten
minutes and learn nothing useful.

The problem is almost never your join logic. It is the strings.

R's `merge()` and dplyr's `*_join()` functions do exactly what you ask: they
compare key values byte-for-byte and link rows that are equal. When keys fail
to match, the join did not make a mistake. The keys are genuinely different.
The difficulty is that the difference is often invisible: a trailing space, a
case mismatch, a zero-width Unicode character that has no visual
representation.

This vignette walks through five scenarios where joins fail for string-level
reasons that resist casual inspection. Each one comes from a real-world
pattern; the data here is synthetic, but the pain is authentic. The goal is
not to catalogue every possible string problem (the `common-issues` vignette
does that), but to show what the diagnosis-to-fix workflow looks like in
practice.

## Scenario 1: The Excel Export

A small retail analytics team receives monthly sales data from a distribution
partner. The partner maintains an Excel workbook; every month they export a CSV
and email it over. The team joins it against their internal customer database
using a shared `customer_id` column.

For two quarters, everything works. The pipeline runs, the row counts check
out, and nobody thinks twice about the join. Then one month, 30% of the sales
records stop matching. Nobody changed the join code. Nobody changed the
customer database. The partner swears the IDs are correct (and they are, in
the sense that every ID in the CSV exists in the customer database). Eyeballing
the CSV confirms that the values look right.

The analyst opens R, loads both tables, and tries `merge()`. The result has
fewer rows than expected. She runs `setdiff()` on the key columns and sees
IDs like `"CUST-1002 "` in the partner data that do not appear in the internal
database. Visually, that looks identical to `"CUST-1002"`. The trailing space
is there, but the console does not render it distinctly.

Here is what the data looks like:

```{r}
partner_sales <- data.frame(
  customer_id = c("CUST-1001", "CUST-1002 ", "CUST-1003",
                  " CUST-1004", "CUST-1005 ", "CUST-1006"),
  amount = c(2500, 1800, 3200, 950, 4100, 1600),
  stringsAsFactors = FALSE
)

internal_db <- data.frame(
  customer_id = c("CUST-1001", "CUST-1002", "CUST-1003",
                  "CUST-1004", "CUST-1005", "CUST-1006", "CUST-1007"),
  region = c("West", "East", "West", "South", "East", "North", "West"),
  stringsAsFactors = FALSE
)
```

Nothing in `str()` reveals the issue. Printing the data frame will not show it
either, because trailing spaces are invisible in console output. But
`join_spy()` catches it immediately:

```{r}
report <- join_spy(partner_sales, internal_db, by = "customer_id")
```

Three of the six partner IDs carry whitespace (a leading space, trailing
spaces, or both). These are distinct character values as far as R is concerned;
`"CUST-1002 "` will never match `"CUST-1002"`.

The fix is one call. `join_repair()` trims whitespace from both tables at once:

```{r}
repaired <- join_repair(partner_sales, internal_db, by = "customer_id")
partner_fixed <- repaired$x
internal_fixed <- repaired$y
```

We can verify the repair worked:

```{r}
key_check(partner_fixed, internal_fixed, by = "customer_id")
```

And now the join gives us what we expected:

```{r}
result <- merge(partner_fixed, internal_fixed, by = "customer_id")
nrow(result)
```

All six partner sales matched. The root cause turned out to be a formula in
the partner's Excel sheet that concatenated values from two columns; one of
those columns had variable-width entries padded with spaces, and the
`CONCATENATE` formula preserved every byte. Nobody noticed because Excel
renders `"CUST-1002"` and `"CUST-1002 "` in exactly the same way. The cell
just shows the text, and trailing whitespace is invisible in a proportional
font.

This is the single most common join failure we have seen. It accounts for
more broken pipelines than all other string issues combined, and it is the
easiest to fix once you know it is there.

## Scenario 2: Two Databases, Two Conventions

Two data sources at a mid-sized SaaS company need to be connected for a
churn analysis. The CRM stores email addresses in uppercase (a decision from
a database migration in the late 1990s that nobody revisited). The web app's
event log stores them in lowercase, because that is what browsers submit. Both
systems have been running independently for years; the team wants to join CRM
profiles to clickstream events to see which subscription plans correlate with
engagement.

```{r}
crm_profiles <- data.frame(
  email = c("ALICE@ACME.COM", "BOB@ACME.COM", "CAROL@ACME.COM",
            "DAVE@ACME.COM", "EVE@ACME.COM"),
  plan = c("enterprise", "starter", "pro", "enterprise", "starter"),
  stringsAsFactors = FALSE
)

click_events <- data.frame(
  email = c("alice@acme.com", "bob@acme.com", "carol@acme.com",
            "dave@acme.com", "frank@acme.com"),
  page_views = c(47, 12, 89, 33, 5),
  stringsAsFactors = FALSE
)
```

The team writes a simple inner join and gets zero rows back. Zero. Not "fewer
than expected": literally nothing. At first they suspect the data did not
load correctly, but both tables have the right number of rows and the columns
are there. The explanation is straightforward: R's string comparison
is case-sensitive, and `"ALICE@ACME.COM"` does not equal `"alice@acme.com"`.
Every single key pair fails to match, even though the underlying email
addresses are obviously the same.

`join_spy()` flags the situation before the join even happens:

```{r}
report <- join_spy(crm_profiles, click_events, by = "email")
```

The case mismatch issue appears in the report. `suggest_repairs()` can generate
the specific code to fix it:

```{r}
suggest_repairs(report)
```

Or we can use `join_repair()` directly, specifying case standardization:

```{r}
repaired <- join_repair(
  crm_profiles, click_events,
  by = "email",
  standardize_case = "lower"
)
```

After lowercasing both sides, the inner join returns four matched rows
(everyone except Eve, who has no click data, and Frank, who is not in the CRM):

```{r}
result <- merge(repaired$x, repaired$y, by = "email")
nrow(result)
result
```

The interesting thing about this scenario is that neither data source is
"wrong." Both are internally consistent; the problem exists only at the
boundary between them. This is the most common shape of join failures in
organizational data work: each system is correct in isolation, maintained by
a different team, documented in its own way. The incompatibility only surfaces
when someone tries to connect them, and the person doing the connecting is
usually not the person who chose the casing convention in either system.

Note that email addresses are case-insensitive by RFC 5321 (the local part
before the @ technically can be case-sensitive, but almost no mail server
enforces this in practice). Lowercasing is the right normalization here. For
other types of identifiers (product codes, country abbreviations, ticker
symbols) you may want `"upper"` instead. The point is to pick one convention
and apply it to both sides before joining.

## Scenario 3: The PDF Copy-Paste

A public health researcher is compiling data from multiple sources for a
systematic review. Most of the data comes from structured databases, but a
handful of studies published their supplementary tables only as PDF. The
researcher needs prevalence estimates from those PDFs, so she selects the
table in the PDF viewer, copies it, pastes it into a spreadsheet, cleans up
the columns, exports to CSV, and reads it into R.

The data looks perfect. Every country name is spelled correctly when she
prints the data frame. The columns have the right types. But when she joins
against a reference table of population data, half the countries fail to
match. She checks for typos manually; there are none. She runs `unique()` on
both country columns and compares them by eye; they look identical.

```{r}
# Simulating PDF copy-paste artifacts:
# \u00A0 is non-breaking space, \u200B is zero-width space
pdf_data <- data.frame(
  country = c("Brazil", "India\u200B", "Germany",
              "Japan\u00A0", "Canada", "France\u200B"),
  prevalence = c(12.3, 8.7, 5.1, 3.9, 6.2, 4.8),
  stringsAsFactors = FALSE
)

reference <- data.frame(
  country = c("Brazil", "India", "Germany", "Japan",
              "Canada", "France", "Italy"),
  population_m = c(214, 1408, 84, 125, 38, 68, 59),
  stringsAsFactors = FALSE
)
```

Watch what happens when we print the PDF data:

```{r}
pdf_data$country
```

Those country names look perfectly clean. The zero-width space after "India"
and "France" is literally invisible; it occupies zero pixels of width. The
non-breaking space after "Japan" renders identically to a regular space (which
`trimws()` would catch), but it is Unicode character U+00A0, not U+0020. A
plain `trimws()` call will not remove it in all R versions.

The merge confirms the damage:

```{r}
nrow(merge(pdf_data, reference, by = "country"))
```

Only three of six countries match. The researcher might spend an hour checking
for typos that do not exist, because the characters causing the failure are
invisible to the human eye.

`join_spy()` detects encoding issues that string inspection cannot:

```{r}
report <- join_spy(pdf_data, reference, by = "country")
```

The report flags invisible characters in the left table. Repair is
straightforward:

```{r}
repaired <- join_repair(pdf_data, reference, by = "country")
nrow(merge(repaired$x, repaired$y, by = "country"))
```

Six matches, as expected. The zero-width spaces and non-breaking spaces are
gone.

This class of problem is hard to catch. Every standard debugging
tool (`print()`, `str()`, `summary()`, `unique()`, even visual comparison
in a spreadsheet) will show you data that looks correct. You can copy the
value from the console, paste it into a string comparison, and it will fail,
and you will not understand why. The characters are there in the byte
representation; they simply have no visible glyph. PDF copy-paste is the most
common source, but web scraping, OCR output, and data from legacy mainframe
systems can produce the same artifacts.

One useful debugging trick outside of joinspy: if you suspect invisible
characters, run `nchar()` on the values. `"India\u200B"` has 6 characters,
not 5. But that requires you to already suspect the problem exists.
`join_spy()` flags it proactively, before you have spent an hour staring at
identical-looking strings.

## Scenario 4: The Slowly Growing Mismatch

Consider a monthly ETL pipeline that has been stable for over a year at an
e-commerce company. Transaction records from the point-of-sale system are joined to
a product catalogue to attach product names, categories, and profit margins.
Finance uses the output for monthly reporting. The pipeline ran cleanly for
months. Then the match rate started drifting downward: 99% in January, 97%
in February, 94% in March. The pipeline did not fail; it just returned
slightly fewer rows each month. Nobody noticed until April, when finance
flagged a discrepancy in the margin totals.

The code had not changed. The product catalogue had not changed. What changed
was a person. A new data entry clerk joined the warehouse team in December.
The existing convention for product codes was strict: `"WDG-100"`, `"WDG-101"`,
uppercase prefix, dash, three-digit numeric suffix. The new clerk sometimes
omitted the dash, sometimes typed in lowercase, sometimes both. The codes
were close enough that the warehouse system accepted them (it did fuzzy
matching internally), but the join in the ETL pipeline did not.

```{r}
# Product catalogue (canonical format)
catalogue <- data.frame(
  product_code = c("WDG-100", "WDG-101", "WDG-102",
                   "WDG-103", "WDG-104", "WDG-105"),
  product_name = c("Widget A", "Widget B", "Widget C",
                   "Widget D", "Widget E", "Widget F"),
  margin = c(0.35, 0.28, 0.42, 0.31, 0.39, 0.25),
  stringsAsFactors = FALSE
)

# Recent transactions (mix of old and new clerk entries)
transactions <- data.frame(
  product_code = c("WDG-100", "WDG-101", "WDG102",
                   "wdg-103", "WDG-104", "wdg105",
                   "WDG-100", "WDG103"),
  quantity = c(5, 3, 7, 2, 4, 6, 1, 8),
  stringsAsFactors = FALSE
)
```

Some of these will match; some will not. The damage is partial, which makes it
harder to notice than a complete failure. `join_spy()` shows the picture:

```{r}
report <- join_spy(transactions, catalogue, by = "product_code")
```

The report flags both case mismatches and unmatched keys. But here is where
this scenario differs from the previous ones: `join_repair()` can fix the case
issue, but it cannot fix the missing dashes. Whitespace trimming and case
standardization are mechanical transformations; inserting a dash into
`"WDG102"` to make it match `"WDG-102"` requires knowledge of the formatting
rule, which the tool does not have.

We can do a dry run to see what `join_repair()` would fix:

```{r}
join_repair(transactions, catalogue,
            by = "product_code",
            standardize_case = "upper",
            dry_run = TRUE)
```

After applying the mechanical fixes:

```{r}
repaired <- join_repair(transactions, catalogue,
                        by = "product_code",
                        standardize_case = "upper")
```

The case issues are resolved, but the dash-versus-no-dash mismatches remain.
Those require a manual transformation:

```{r}
# Manual fix: insert dash if missing in product codes matching the pattern
fix_codes <- function(codes) {
  gsub("^([A-Z]{3})(\\d)", "\\1-\\2", codes)
}
repaired$x$product_code <- fix_codes(repaired$x$product_code)
```

Now we can verify:

```{r}
result <- merge(repaired$x, repaired$y, by = "product_code")
nrow(result)
```

All eight transactions match. This scenario is worth sitting with for a
moment, because it illustrates a boundary in what automated repair can do.
`join_repair()` handles mechanical transformations: trimming whitespace,
normalizing case, stripping invisible characters. These are context-free
operations that are always safe to apply. But inserting a dash into `"WDG102"`
to produce `"WDG-102"` requires knowing that the canonical format uses a dash
after three letters. That is domain knowledge. `join_spy()` will show you that
keys are unmatched and flag near-matches, but the fix has to come from someone
who understands the data.

In practice, the right response to this kind of drift is not just a regex in
the pipeline; it is a conversation with the warehouse team about enforcing
the code format at the point of entry. The pipeline fix is a bandage; the
process fix is the cure.

## Scenario 5: The Compound Key Trap

This one is subtle. Two government datasets need to be linked for a policy
briefing: regional economic indicators and regional population estimates.
Both tables are keyed on region and year, a compound key with two columns.
The year column is numeric, so it matches without trouble. The region column
is character, and it has a problem that affects only some of the records.

```{r}
economics <- data.frame(
  region = c("North America", "Europe", "Asia Pacific ",
             "North America", "Europe", "Asia Pacific "),
  year = c(2022, 2022, 2022, 2023, 2023, 2023),
  gdp_growth = c(2.1, 1.8, 4.2, 1.9, 0.9, 3.8),
  stringsAsFactors = FALSE
)

population <- data.frame(
  region = c("North America", "Europe", "Asia Pacific",
             "North America", "Europe", "Asia Pacific"),
  year = c(2022, 2022, 2022, 2023, 2023, 2023),
  pop_millions = c(580, 450, 4300, 585, 448, 4350),
  stringsAsFactors = FALSE
)
```

In a compound key join, both columns must match for a row to link. The year
column is fine everywhere. But `"Asia Pacific "` (with a trailing space) in the
economics table will not match `"Asia Pacific"` in the population table. The
result: North America and Europe link correctly; Asia Pacific silently
vanishes.

```{r}
merged <- merge(economics, population, by = c("region", "year"))
nrow(merged)
```

Four rows instead of six. Two-thirds of the data linked successfully, which is
just enough to look plausible if you are not paying close attention. The
analyst producing the policy briefing might not notice that Asia Pacific is
missing entirely, especially if the briefing template just shows a table of
whatever regions came through, without verifying completeness. Someone running
this pipeline on autopilot might not catch the drop for weeks, and when they
do notice, they might blame the data provider rather than the string encoding.

`join_spy()` pinpoints which column in the compound key has the problem:

```{r}
report <- join_spy(economics, population, by = c("region", "year"))
```

The whitespace issue is flagged specifically in the `region` column. We repair
just that column and leave `year` untouched (it is numeric; there is nothing
to trim):

```{r}
repaired <- join_repair(economics, population, by = c("region", "year"))
result <- merge(repaired$x, repaired$y, by = c("region", "year"))
nrow(result)
```

Six rows. All regions, both years. The fix was trivial once the problem was
identified; finding the problem was the hard part.

Compound keys are more fragile than single keys precisely because the failure
mode is partial. A single-column key with whitespace issues will drop rows in
a way that is at least dramatic enough to investigate. A compound key with one
clean column and one dirty column will match on some rows and fail on others,
producing a result that looks reasonable but is quietly incomplete. The more
columns in the key, the more places a single invisible character can hide.

## The Pattern

Every scenario in this vignette shares three properties.

First, the data looks correct to standard inspection tools. `str()` shows
character columns of the right type. `summary()` reports the right number of
unique values (give or take). `print()` renders the values identically. Nothing
in the usual diagnostic workflow reveals the problem. The tools are not broken;
they were simply not designed for this. `str()` tells you the type and length
of a vector. It does not tell you whether byte 7 of the third element is a
regular space or a non-breaking space.

Second, the failure is silent. R does not warn you when a merge drops rows
because of invisible whitespace; it just returns fewer rows. If you are not
checking row counts before and after every join (and many pipelines do not),
the error propagates downstream without any signal that something went wrong.
A left join will still return all rows from the left table, but the columns
from the right table will be filled with `NA` where the match failed, and
those NAs might look like legitimate missing data.

Third, the fix is mechanical once you know the cause. Trimming whitespace,
lowercasing strings, stripping invisible Unicode: these are one-line
operations. The repair is trivial; the diagnosis is where all the time goes. An analyst
who knows that the problem is trailing whitespace can fix it in thirty seconds.
An analyst who does not know that will spend hours comparing values that look
identical, questioning the join logic, rebuilding the pipeline from scratch,
and still not finding the issue.

`join_spy()` was built to close that gap between the thirty-second fix and the
hours of confusion that precede it. Run it before your joins, especially when
working with data from external sources, manual entry, PDF extraction, or
cross-system integrations. The function checks for the problems that `str()`
and `summary()` were never designed to catch.
